# Running R from the command line {#sec-rscript}

```{r}
#| include: false
source("_common.R")
```

Snakemake does not run R code. It runs *commands*, and it decides when to run
them. So before any of it is useful, your analysis has to be something you can
start from a terminal without clicking anything --- and, ideally, something you
can point at different parameter values without editing the file.

This chapter has nothing to do with Snakemake. If you already write scripts that
take arguments and save their output to a file you name, skip to
@sec-first-snakefile.

## Interactive work is not reproducible

It is worth being blunt about this once, because everything else in the book
follows from it.

Whatever is in your R session right now is the product of a history nobody has,
including you. Objects created by code you have since edited or deleted. A package
loaded by hand in March. A seed set
before an experiment you abandoned. And if you saved the workspace on
exit, all of that invisible state came back with you the next morning.

The practical consequence is that "it works" is not a claim you can check. A number
in your session may or may not be what the current version of your code produces.
There is no way to find out except to start a clean session and run the code from
the top --- which is exactly what running a script does.

None of this is an argument against working interactively. Interactive is how you
figure out what the code should be; it is the right tool for exploring a dataset,
staring at a model, and getting a plot to look right. The rule is narrower than
that: **nothing may exist only in your session.** Once a piece of analysis is real,
it belongs in a file that runs top to bottom, in a fresh session, and writes its
output somewhere you named.

::: {.callout-tip}
## Don't save your workspace

Turn off the `.RData` prompt --- in RStudio, *Tools → Global Options → General*,
set "Save workspace to .RData on exit" to *Never* and clear "Restore .RData into
workspace at startup". Per project, the same two settings are `SaveWorkspace: No`
and `RestoreWorkspace: No` in the `.Rproj` file; this repository's has both.

It feels like losing something. What you lose is the illusion that your results
survive a restart, and finding that out early is much cheaper than finding it out
the week a paper is due.
:::

Snakemake is the machinery that makes the rule livable: it reruns from a clean
session only the parts that are actually out of date, so insisting on scripts stops
costing you the thing interactive work was buying.

## Rscript

`Rscript` runs an R file top to bottom and exits:

```bash
Rscript sim.R
```

Two things change when your code runs this way rather than in RStudio.

The first is that the working directory is wherever your terminal is, not wherever
the file is. If `sim.R` says `readRDS("results/thing.rds")`, that path is resolved
relative to the directory you were sitting in when you typed the command. Pick a
convention and hold to it; the one used throughout this book is **always run from
the project root**, so every path in every script is relative to the top of the
project. Snakemake follows the same convention by default, which is convenient.

The second is that nothing prints unless you ask. Auto-printing at the top level
is an interactive feature, so `head(d)` on its own line produces no output under
`Rscript`; you need `print(head(d))` or `cat()`. This matters more than it sounds
like it should, because those prints become your log file in @sec-parallel.

## Taking arguments

A script that hard-codes `n <- 100` can only ever do one thing. To run the same
code at three sample sizes you would have to edit the file between runs, which is
exactly the kind of manual step we are trying to eliminate. Instead, read the
values from the command line.

The base R way is `commandArgs()`, which hands you the arguments as a character
vector:

```{r}
#| echo: false
#| output: asis
embed_file("examples/01-first-rules/sim.R", from = "^args <- commandArgs",
           to = "^outfile", filename = "sim.R (excerpt)")
```

and is then called like this:

```bash
Rscript sim.R naive results/sim-naive.rds
```

This is fine for two arguments. It gets bad quickly: the arguments are positional,
so you have to remember the order, everything arrives as a character string that
you have to convert yourself, and there is no way to give something a default.

## optparse

The [`optparse`](https://cran.r-project.org/package=optparse) package gives you
named flags with types and defaults, which is what the rest of this book uses:

```{r}
#| echo: false
#| output: asis
embed_file("examples/02-wildcards/sim.R", from = "^library\\(optparse\\)",
           to = "^opt <- parse_args", filename = "sim.R (excerpt)")
```

Now the script is called with named arguments in any order:

```bash
Rscript sim.R --n 400 --rho 80 --method split --out results/sim/n400_rho80_split.rds
```

and `opt` is a list: `opt$n` is an integer, `opt$method` is a string. Anything you
leave off falls back to its default, so the script still runs as-is when you are
poking at it interactively.

::: {.callout-tip}
## Defaults are documentation

Set the defaults to the values you use most often. Someone reading
`make_option("--iterations", type = "integer", default = 200)` learns both what
the flag is called and what a normal value looks like, and you get a script that
runs unmodified when you source it by hand.
:::

## One script, one output file

The last convention, and the one Snakemake actually requires: **a script should
write to a path it was told, not a path it invented.**

```{r}
#| echo: false
#| output: asis
embed_file("examples/02-wildcards/sim.R", from = "^dir.create",
           to = "^cat\\(\"wrote", filename = "sim.R (excerpt)")
```

Note the `dir.create()`. Snakemake creates the directories for a rule's declared
outputs before running it, but scripts that also work when you run them by hand
shouldn't count on that.

This is what makes a script composable. `sim.R` does not know it is part of a
pipeline; it knows how to turn a set of parameters into one file. The Snakefile in
the next chapter is nothing more than a list of which parameters go with which
filename.

## Where the results go

Two decisions that will save you pain later:

**Results are files, and files are named for what produced them.** A file called
`n400_rho80_split.rds` tells you the sample size, the correlation, and the method.
A file called `results2_final.rds` tells you nothing, and in six months neither
will you.

**Results are regenerable, so they don't go in git.** Commit the code and the
inputs; let the outputs be rebuilt.

**Everything a project computes goes in one directory.** Every example in this
book has a `results/` folder at its root, and nothing is written anywhere else.
One directory to ignore in git, one to delete when you want to prove the pipeline
still works, one to hand to a collaborator. @sec-shape says more about the layout
once you have seen a Snakefile.
