# Running R from the command line {#sec-rscript}

```{r}
#| include: false
source("_common.R")
```

Snakemake does not run R code. It runs *commands*, and it decides when to run
them. So before any of it is useful, your analysis has to be something you can
start from a terminal without clicking anything --- and, ideally, something you
can point at different parameter values without editing the file.

This chapter has nothing to do with Snakemake. If you already write scripts that
take arguments and save their output to a file you name, skip to
@sec-first-snakefile.

## Rscript

`Rscript` runs an R file top to bottom and exits:

```bash
Rscript sim.R
```

Two things change when your code runs this way rather than in RStudio.

The first is that the working directory is wherever your terminal is, not wherever
the file is. If `sim.R` says `readRDS("out/thing.rds")`, that path is resolved
relative to the directory you were sitting in when you typed the command. Pick a
convention and hold to it; the one used throughout this book is **always run from
the project root**, so every path in every script is relative to the top of the
project. Snakemake follows the same convention by default.

The second is that nothing prints unless you ask. Auto-printing at the top level
is an interactive feature, so `head(d)` on its own line produces no output under
`Rscript`; you need `print(head(d))` or `cat()`. This matters more than it sounds
like it should, because those prints become your log file in @sec-parallel.

## Taking arguments

A script that hard-codes `n <- 100` can only ever do one thing. To run the same
code at three sample sizes you would have to edit the file between runs, which is
exactly the kind of manual step we are trying to eliminate. Instead, read the
values from the command line.

The base R way is `commandArgs()`, which hands you the arguments as a character
vector:

```{r}
#| echo: false
#| output: asis
embed_file("examples/01-first-rules/sim.R", from = "^args <- commandArgs",
           to = "^outfile", filename = "sim.R (excerpt)")
```

and is then called like this:

```bash
Rscript sim.R naive ../../results/01-first-rules/sim-naive.rds
```

This is fine for two arguments. It gets bad quickly: the arguments are positional,
so you have to remember the order, everything arrives as a character string that
you have to convert yourself, and there is no way to give something a default.

## optparse

The [`optparse`](https://cran.r-project.org/package=optparse) package gives you
named flags with types and defaults, which is what the rest of this book uses:

```{r}
#| echo: false
#| output: asis
embed_file("examples/02-wildcards/sim.R", from = "^library\\(optparse\\)",
           to = "^opt <- parse_args", filename = "sim.R (excerpt)")
```

Now the script is called with named arguments in any order:

```bash
Rscript sim.R --n 400 --rho 80 --method split \
  --out ../../results/02-wildcards/sim/n400_rho80_split.rds
```

and `opt` is a list: `opt$n` is an integer, `opt$method` is a string. Anything you
leave off falls back to its default, so the script still runs as-is when you are
poking at it interactively.

::: {.callout-tip}
## Defaults are documentation

Set the defaults to the values you use most often. Someone reading
`make_option("--iterations", type = "integer", default = 200)` learns both what
the flag is called and what a normal value looks like, and you get a script that
runs unmodified when you source it by hand.
:::

## One script, one output file

The last convention, and the one Snakemake actually requires: **a script should
write to a path it was told, not a path it invented.**

```{r}
#| echo: false
#| output: asis
embed_file("examples/02-wildcards/sim.R", from = "^dir.create",
           to = "^cat\\(\"wrote", filename = "sim.R (excerpt)")
```

Note the `dir.create()`. Snakemake creates the directories for a rule's declared
outputs before running it, but scripts that also work when you run them by hand
shouldn't count on that.

This is what makes a script composable. `sim.R` does not know it is part of a
pipeline; it knows how to turn a set of parameters into one file. The Snakefile in
the next chapter is nothing more than a list of which parameters go with which
filename.

## Where the results go

Three decisions that will save you pain later:

**Results are files, and files are named for what produced them.** A file called
`n400_rho80_split.rds` tells you the sample size, the correlation, and the method.
A file called `results2_final.rds` tells you nothing.

**Results are regenerable, so they don't go in git.** Commit the code and the
inputs; let the outputs be rebuilt --- which is a claim you can only make honestly
once the pipeline is complete.

**All the results go in one place.** Every example in this book writes its
simulation output to a single shared folder, `results/`, at the top of the
repository:

```{.default filename="the layout"}
snakemake-tutorial/
  results/                 <- every example's computed results
    01-first-rules/
    02-wildcards/
    ...
  examples/
    01-first-rules/
      sim.R  figure.R  Snakefile
      out/                 <- figures and tables, rebuilt in seconds
```

The split is between things that are expensive to compute and things that are
derived from them. The `.rds` files are the expensive part, so they go in the
shared folder; a figure that takes two seconds to redraw stays next to the code
that draws it, in `out/`.

Why bother, when each example could keep its own results next door? Because one
folder is a thing you can hand to somebody. Back it up, sync it, put it on a
shared drive, and a collaborator has every result the project has ever computed
without also getting your working directory. @sec-scaling comes back to this once
the pipeline is big enough for it to matter.
