# A first Snakefile {#sec-first-snakefile}

```{r}
#| include: false
source("_common.R")
```

Start in `examples/01-first-rules`. It contains two R scripts --- `sim.R`, which
runs the simulation for one method, and `figure.R`, which turns simulation output
into a figure --- and a file called `Snakefile`.

By hand, building the figure means running three commands in the right order:

```bash
Rscript sim.R naive rds/sim-naive.rds
Rscript sim.R split rds/sim-split.rds
Rscript figure.R out/figure1.pdf rds/sim-naive.rds rds/sim-split.rds
```

That is not hard. It is just something you have to remember, and remember *again*
every time you change one of the scripts, and remember correctly --- because
nothing stops you from rebuilding the figure without rerunning the simulation.

## Rules

A Snakefile is a list of rules. A rule says: here are the files I need, here is the
file I produce, and here is the command that gets from one to the other.

```{r}
#| echo: false
#| output: asis
embed_file("examples/01-first-rules/Snakefile")
```

Read `rule sim_naive` as a sentence: *to make `rds/sim-naive.rds`, given
`sim.R`, run `Rscript sim.R naive rds/sim-naive.rds`.*

A few mechanics:

- **`{output}` and `{input}` are substituted into the shell command.** You write
  the filename once, in the `output:` block, and refer to it after that. This is
  not just brevity --- it makes it impossible for the rule to declare one file and
  write a different one.
- **Inputs can be named**, as in `rule figure`, where `input.script` and
  `input.rds` are referred to separately in the command. Without names, `{input}`
  expands to all of them, space-separated, in order.
- **Indentation matters**, as in Python. Use four spaces and stay consistent.
- **`sim.R` is listed as an input**, even though it is code rather than data. This
  is important. It is what makes "the script changed" a reason to rerun.

## rule all

`rule all` is different. It has no output and no command --- only inputs. It exists
because of a rule Snakemake follows strictly:

> Snakemake builds the first rule in the file, and whatever that rule needs.

So `rule all` is where you list the things you actually want to exist at the end.
Everything else gets built because something in that list, directly or indirectly,
depends on it. A file that is not reachable from `rule all` will not be built, no
matter how many rules could make it. By convention this is named `rule all`, but what matters is that it is first.

## The dependency graph

From those four rules Snakemake works out the order for itself. Nothing in the
file says the simulations come before the figure; it deduces that from the fact
that `rule figure` needs files that `rule sim_naive` and `rule sim_split` produce.

```{mermaid}
%%| fig-align: center
flowchart LR
  A[sim.R] --> B[rds/sim-naive.rds]
  A --> C[rds/sim-split.rds]
  D[figure.R] --> E[out/figure1.pdf]
  B --> E
  C --> E
  E --> F([rule all])
```

This is a directed acyclic graph (DAG). Two consequences
worth internalizing early: anything with no path between the nodes can run at the
same time (@sec-parallel), and a cycle is an error rather than an infinite loop (it won't run).

## Running it

Always look before you leap. `-n` (or `--dry-run`) reports what *would* happen:

```{r}
#| echo: false
#| output: asis
transcript("01-dry-run.txt")
```

Note the `reason:` line on each job --- Snakemake tells you not just what it plans
to do but why. Here everything is missing, so everything runs.

Now do it, allowing two jobs at a time:

```{r}
#| echo: false
#| output: asis
transcript("01-run.txt")
```

`--cores` is not optional; Snakemake refuses to run without being told how much of
your machine it may use. The two simulations ran concurrently, the figure waited
for both, and the whole thing took about four seconds.

## Doing nothing, quickly

Run it again:

```{r}
#| echo: false
#| output: asis
transcript("01-up-to-date.txt")
```

This is the point of the exercise. Ask for the figure a hundred more times and you
will get that message a hundred times. Now delete the figure and ask again:

```{r}
#| echo: false
#| output: asis
transcript("01-rebuild-figure.txt")
```

Only the figure was rebuilt. The simulations were still there and still current, so
they were left alone --- Snakemake rebuilds what is out of date and nothing else.

## What counts as out of date

Here is where Snakemake differs from `make`, and from most people's expectations.

`make` asks one question: is the input newer than the output? If so, rebuild.

Snakemake asks two, and needs both answers to be yes. It keeps a record of every
output it has ever made --- in a directory called `.snakemake/` --- including a
SHA-256 checksum of each input that went into it. A job is out of date when

1. an input's timestamp is newer than the output's, **and**
2. that input's contents differ from what was recorded.

The timestamp is the gate and the checksum is the confirmation. So:

```{r}
#| echo: false
#| output: asis
transcript("04-touch-is-not-enough.txt")
```

Touching a file does not trigger a rebuild, because the contents did not change.
Change one character of `sim.R`, though, and everything downstream of it goes:

```{r}
#| echo: false
#| output: asis
transcript("04-cascade.txt")
```

(That transcript comes from the larger example of @sec-manuscript, which is why
there are 23 jobs rather than 4.)

Content is not the only trigger. Snakemake also stores the shell command and the
parameters used for each output, so editing a rule's command --- not just the
script it calls --- makes the affected outputs stale too. @sec-debugging covers how
to inspect and override all of this.

::: {.callout-warning}
## The timestamp really is a gate

Because the check is an *and*, a changed file whose timestamp is not newer than
the output is invisible: Snakemake never gets as far as comparing the checksum. You
will not produce that situation by editing a file --- editing sets the timestamp to
now --- but you can arrive at it by other routes, because several common tools
preserve timestamps deliberately:

```bash
cp -p old/sim.R code/sim.R      # -p preserves the original mtime
rsync -a backup/ .              # -a implies -t
tar xf archive.tar              # restores the archived mtimes
```

Restore a script from a backup or an archive that way and the pipeline can sit
there insisting everything is up to date. If you have moved files around by any
means other than editing them, `touch` the ones you restored, or force the issue
with `snakemake -R` (@sec-debugging).
:::

::: {.callout-note}
## Deleting `.snakemake/` is safe

It is a cache of metadata, not results. Delete it and the next run rebuilds
everything, because Snakemake no longer has any record of how the existing files
were made. Add it to `.gitignore` and otherwise ignore it.
:::

## Building one thing

You can also name a target directly, which is invaluable when you are developing
and want to check one piece:

```bash
snakemake --cores 1 rds/sim-naive.rds
```

Snakemake finds the rule that can produce that file, builds whatever it needs, and
stops. You can also name a *rule* --- `snakemake --cores 1 figure` --- which asks
for that rule's outputs.

## One Snakefile, one project {#sec-shape}

Before going further, it is worth being explicit about what a "project" is here,
because every example in this book is one and your own work should be too.

`examples/01-first-rules` is a complete, self-contained project. Copy that
directory anywhere on your disk and everything still
works --- it has no dependency on the rest of this repository. The same is true of
each of the other four. They are laid out side by side for convenience, but each is
a miniature of a real repository: one paper, one analysis, one thing.

The layout that scales from these examples up to a real methods paper is:

```{.default filename="the shape of a project"}
my-project/
  Snakefile          <- the map: every rule, in one file, at the root
  config.yaml        <- the settings (chapter 4)
  code/              <- the scripts the rules call
  rds/               <- computed results
  out/               <- figures and tables
```

Three of those deserve to be stated as rules rather than suggestions.

**One Snakefile, at the root.** Not one per subdirectory. The Snakefile is the
only place that describes how the pieces of the project fit together, so a
newcomer --- a coauthor, a reviewer, you in a year --- has exactly one file to
open to find out what this project builds and what it builds it from. Snakemake
does let you split rules across files with `include:`, and that is worth reaching
for once a single file becomes genuinely unwieldy, but the default should be one
file you can read top to bottom.

**`rds/` holds what the pipeline computes.** Simulation output, fitted models,
cleaned data --- the objects that took time to produce and that everything else is
derived from. They are serialized R objects, hence the name; use whatever name you
like, but use one directory.

**`out/` holds what a human looks at.** Figures and tables: the things that end up
in the manuscript or the report.

The reason to split these rather than pile everything into one `results/` folder is
that the two halves behave completely differently. `rds/` is expensive, is reused
across a paper and a talk and a grant renewal, and may take days to regenerate ---
it is the thing worth backing up and worth sending to a coauthor. `out/` is cheap:
delete it and `snakemake` puts it back in seconds, from the contents of `rds/`.
Once they are separate directories you can treat them differently, and
@sec-scaling does exactly that.

The corollary is that anything *not* in `rds/` or `out/` is either code, settings,
or something a human wrote. That is a useful thing to be able to say about a
directory --- and it is what makes the next rule easy to apply.

**Neither `rds/` nor `out/` is tracked by git.** Both are outputs. Git holds the
things a person wrote --- the scripts, the Snakefile, the config, the `.tex` files
--- and nothing that a program produced:

```{.default filename=".gitignore"}
rds/
out/
.snakemake/
```

`rds/` is the easy case; nobody wants four hundred megabytes of serialized objects
in a repository. `out/` is the one people talk themselves out of, because figures
are small and it is genuinely convenient to look at a PDF on GitHub. Do it anyway.
A committed figure is a second copy of a result, and the moment it exists it can
disagree with the code that made it --- which is the exact failure this book is
about. If you want to see the figures, run `snakemake`; it takes seconds, and what
you get is guaranteed to match the repository you are looking at.

The one thing this costs you is that a fresh clone cannot show anybody a result
until it has been rebuilt. When the rebuild takes days rather than seconds that is
a real cost, and @sec-scaling deals with it --- by sharing `rds/` outside of git,
not by committing it.

## Try it

1. Delete `out/` and rebuild. Time it. Then delete `rds/` as well and compare.
2. Change the `n <- 100` line in `sim.R` to `n <- 200` and run `snakemake -n`.
   Which jobs are planned, and why?
3. Add a line to `rule all` asking for `out/nonexistent.pdf` and run
   `snakemake -n`. Read the error carefully --- you will see it again.
