# Rebuilding only what changed {#sec-rerun}

```{r}
#| include: false
source("_common.R")
```

@sec-first-snakefile ended with a working pipeline. What makes one worth having is
not that it runs, but what it does the *second* time you ask --- and, once a
project is real, what it decides to leave alone.

## Doing nothing, quickly

Still in `examples/01-first-rules`, run it again:

```{r}
#| echo: false
#| output: asis
transcript("01-up-to-date.txt")
```

This is the point of the exercise. Ask for the figure a hundred more times and you
will get that message a hundred times. Now delete the figure and ask again:

```{r}
#| echo: false
#| output: asis
transcript("01-rebuild-figure.txt")
```

Only the figure was rebuilt. The simulations were still there and still current, so
they were left alone --- Snakemake rebuilds what is out of date and nothing else.

## What counts as out of date

Here is where Snakemake differs from `make`, and from most people's expectations.

`make` asks one question: is the input newer than the output? If so, rebuild.

Snakemake asks two, and needs both answers to be yes. It keeps a record of every
output it has ever made --- in a directory called `.snakemake/` --- including a
SHA-256 checksum of each input that went into it. A job is out of date when

1. an input's timestamp is newer than the output's, **and**
2. that input's contents differ from what was recorded.

The timestamp is the gate and the checksum is the confirmation. So:

```{r}
#| echo: false
#| output: asis
transcript("04-touch-is-not-enough.txt")
```

Touching a file does not trigger a rebuild, because the contents did not change.
Change one character of `sim.R`, though, and everything downstream of it goes:

```{r}
#| echo: false
#| output: asis
transcript("04-cascade.txt")
```

(That transcript comes from the larger example of @sec-manuscript, which is why
there are 22 jobs rather than 4.)

Content is not the only trigger. Snakemake also stores the shell command used for each output, so editing a rule's command --- not just the
script it calls --- makes the affected outputs stale too. 

@sec-debugging covers how
to inspect and override all of this.


## Declare what a script sources

Scripts grow to the point where they share things --- a set of functions, the
labels and colors several figures use --- and those live in files of their own,
which is what `R/` is for in @sec-shape for shared functions. Snakemake has no way to know that a script
reads another file unless you tell it, so a shared file has to be listed as an input to every rule
whose script sources it. An undeclared dependency is the one failure mode Snakemake
cannot warn you about: edit a shared function and every result built from it stays
quietly stale.

That is also an argument for keeping such files narrow, because the declaration is
all-or-nothing. In the larger example of @sec-manuscript the simulation functions
are in `R/` and the plot labels and colors are in `scripts/setup.R`, listed
separately: editing a hex code plans four jobs and leaves the simulations alone,
while editing a function plans all twenty-two. Had both lived in one file, the
color change would have rerun eighteen simulations, because Snakemake cannot tell
which half of a file you touched.

## Try it

1. Delete `out/` and rebuild. Time it. Then delete `rds/` as well and compare.
2. Change the `n <- 100` line in `sim.R` to `n <- 200` and run `snakemake -n`.
   Which jobs are planned, and why?
3. Change a comment in `figure.R`, then run `snakemake -n`. Does the simulation
   rerun? Should it?
