# Configuration {#sec-config}

```{r}
#| include: false
source("_common.R")
```

`sim.R` in @sec-wildcards has 200 iterations written into it, which is the wrong
place: you will want 20 while you are debugging and 1000 for the final results, and
changing that should not mean editing the simulation code.

Move to `examples/03-config`, which puts all of it in one file.

## config.yaml

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/config.yaml")
```

YAML is about as simple as it looks: `key: value`, one per line, lists in square
brackets, `#` for comments. Load it at the top of the Snakefile:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^configfile:", to = "^configfile:",
           filename = "Snakefile (excerpt)")
```

and everything in it is available as `config`, a dictionary: `config["iterations"]`
is `200`. Use `config.get("key", default)` when a setting is optional.

Notice how little is in there. A config is for settings that apply to the project
as a whole --- `iterations` means the same thing to every rule that computes
anything.

## The config is an input {#sec-config-input}

One more line is worth adding before going further. `config.yaml` decides what the
numbers are, so anything built from those numbers depends on it --- and saying so
is a matter of listing it, exactly like a script:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^rule figure:", to = "^        config = ",
           filename = "Snakefile (excerpt)")
```

Now the figure carries a recorded dependency on the settings that produced it. Edit
`config.yaml` --- even to add a comment --- and Snakemake reports `Updated input
files: config.yaml` and rebuilds it, for the same reason it does when `figure.R`
changes: the contents no longer match what was recorded.

Note where the line is *not* listed. `rule sim` does not take `config.yaml` as an
input, deliberately, because otherwise any edit to the config would invalidate
every simulation on disk --- and in a real project that is days of compute thrown
away because somebody reworded a comment. Put it on the rules that assemble your
products, not on the ones that do the expensive work.

This matters most at the far end of a pipeline. @sec-manuscript lists `config.yaml`
among the manuscript's inputs, so the compiled PDF is tied to the configuration
that produced the numbers in it; the real project in @sec-scaling does the same.
Commit the config, and the paper you built last March can be rebuilt from the tag.

## Runs that don't overwrite each other {#sec-iterations-path}

This one setting is treated differently, and it is the most useful idea in this chapter.
`iterations` is not passed to the script and forgotten --- it names the directory
the results go in:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^## The iteration count",
           to = "^ITER", filename = "Snakefile (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^rule sim:", to = '^        "--out',
           filename = "Snakefile (excerpt)")
```

`{ITER}` never appears in that rule: the count arrives as the wildcard
`{wildcards.iterations}`, read back out of the filename that was asked for by `rule figure`, just
like `{wildcards.n}`.

The figure rule collects the set `ITER` points at:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "        rds = expand",
           to = "                     method = ", filename = "Snakefile (excerpt)")
```

## wildcard_constraints

@sec-wildcards warned that a wildcard matches anything. Constraints fix that:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^wildcard_constraints:",
           to = '^    method = ', filename = "Snakefile (excerpt)")
```

`r"\d+"` is a regular expression matching one or more digits; the `r` prefix marks a
raw string, so the backslash reaches the regex intact. Declared at the top level
they apply to every rule, and they can also be attached to a single rule, as
@sec-manuscript does.

Two reasons to bother:

**Ambiguity.** With unconstrained wildcards there is more than one way to match
`n{n}_rho{rho}_{method}.rds` against a name containing extra underscores, and
Snakemake resolves it by its rules rather than yours.

**Failing early.** Without the `\d+` on `{n}`, asking for
`rds/200/sim/nfifty_rho0_naive.rds` matches quite happily ---

```{.default}
wildcards: iterations=200, n=fifty, rho=0, method=naive
```

--- and Snakemake goes off and runs `sim.R --n fifty`, which dies inside R however
long the job takes to get that far. With the constraint, nothing can produce that
file and you are told so before anything starts.

A misspelled *extension* is a different matter: `n50_rho0_naive.RDS` fails whether
or not you have constraints, because the pattern ends in a literal `.rds` and
`.RDS` is not that. Constraints police the wildcards, not the text around them.

## Overriding from the command line

Any config value can be replaced for a single run with `--config`:

```bash
snakemake --cores 4 --config iterations=20
```

which is the fastest way to check that a pipeline is wired up before committing to
the real thing:

```{r}
#| echo: false
#| output: asis
transcript("03-smoke-vs-full.txt")
```

Five seconds against eleven, for a grid that would take considerably longer at the
iteration counts of a real paper.

::: {.callout-warning}
## `--config` does not always rebuild what you expect

What an override actually rebuilds depends on how the setting is used, and the two
cases here behave differently:

- **`--config iterations=20`** asks for *different files*. Eighteen new simulations
  run; the old ones are left alone.
- **Dropping a sample size from a rule's `expand()`** is a Snakefile edit rather
  than a config one, and it reruns no simulations at all --- the remaining results
  are still perfectly valid. The figure does rerun, with the reason `Set of input
  files has changed since last execution`, because it is now summarizing twelve
  files instead of eighteen.

Note that `--config` wants one `key=value` per setting and will not take a list on
the command line (`--config n=[50,100]` is a parse error on 9.13). Change the grid
in `config.yaml`, or point `--configfile` at a second YAML file whose values
override the first.

When you want to force the issue, `snakemake --forcerun sim` reruns every job of
that rule regardless. See @sec-debugging.
:::

## Where to draw the line

A useful test: **if changing it changes the results, it belongs in the config; if
changing it changes what the code does, it belongs in the code.**

Iteration counts and file locations --- config. The definition of the estimator ---
code. The temptation, once you have a config file, is to make everything
configurable, and you end up with a 200-line YAML file harder to read than the code
it configures. The config here has one setting, and it is the one that genuinely
applies to everything.

## Try it

1. Run with `--config iterations=20`, then again with `--config iterations=200`.
   What does Snakemake rerun, and why?
2. Add `1000` to the `n` list in `config.yaml` and run `snakemake -n`.
3. Delete the `wildcard_constraints` block and ask for
   `rds/sim/n50_rho0_na_ive.rds`. What happens, and what did `{method}` match?
