# Configuration {#sec-config}

```{r}
#| include: false
source("_common.R")
```

The Snakefile in @sec-wildcards has the parameter grid written into it, and `sim.R`
has 200 iterations and the seed written into it. Both are the wrong places. You will
want 20 iterations while you are debugging and 1000 for the final results, and a
coauthor will want to add a sample size without reading your Python.

Move to `examples/03-config`, which puts all of it in one file.

## config.yaml

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/config.yaml")
```

YAML is about as simple as it looks: `key: value`, one per line, lists in square
brackets, `#` for comments. Load it at the top of the Snakefile:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^configfile:", to = "^configfile:",
           filename = "Snakefile (excerpt)")
```

and everything in it is available as `config`, a dictionary --- `config["iterations"]`
is `200` and `config["n"]` is the list `[50, 100, 400]`. Use `config.get("key",
default)` when a setting is optional.

## params

Values from the config reach a shell command through `params`:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^rule sim:", to = '^        "--out',
           filename = "Snakefile (excerpt)")
```

You could write `config["iterations"]` straight into the command with a Python
f-string and it would work. Declaring it in `params` is better for one specific
reason: Snakemake records the params it used alongside each output, so changing a
parameter marks the affected files stale, exactly as changing an input file does.
Values smuggled into the command another way are invisible to that machinery.

The grid comes from the config now too, so `expand()` reads:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "        rds = expand",
           to = "                     n=config", filename = "Snakefile (excerpt)")
```

Adding a fourth sample size is now a two-character edit to `config.yaml`.

## wildcard_constraints

@sec-wildcards warned that a wildcard matches anything. Constraints fix that:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^wildcard_constraints:",
           to = '^    method = ', filename = "Snakefile (excerpt)")
```

`r"\d+"` is a regular expression matching one or more digits; the `r` prefix marks a
raw string, so the backslash reaches the regex intact. Declared at the top level
they apply to every rule, and they can also be attached to a single rule, as
@sec-manuscript does.

Two reasons to bother:

**Ambiguity.** With unconstrained wildcards there is more than one way to match
`n{n}_rho{rho}_{method}.rds` against a name containing extra underscores, and
Snakemake resolves it by its rules rather than yours.

**Failing early.** A typo like `n50_rho0_naive.RDS` now fails immediately with "no
rule to produce" instead of launching a job that dies twenty minutes later inside R.

## Overriding from the command line

Any config value can be replaced for a single run with `--config`:

```bash
snakemake --cores 4 --config iterations=20
```

which is the fastest way to check that a pipeline is wired up before committing to
the real thing:

```{r}
#| echo: false
#| output: asis
transcript("03-smoke-vs-full.txt")
```

Five seconds against eleven, for a grid that would take considerably longer at the
iteration counts of a real paper.

::: {.callout-warning}
## `--config` does not always rebuild what you expect

`--config iterations=20` changes a `params` value, so the affected outputs are
marked stale and rerun. A config value used only to build the list inside
`expand()` --- the grid itself --- behaves differently: dropping a sample size just
means fewer files are asked for, and the files still being asked for are still
valid, so nothing reruns. That is correct, and occasionally surprising.

When you want to force the issue, `snakemake --forcerun sim` reruns every job of
that rule regardless. See @sec-debugging.
:::

## Where to draw the line

A useful test: **if changing it changes the results, it belongs in the config; if
changing it changes what the code does, it belongs in the code.**

Iteration counts, seeds, grids, confidence levels, file locations --- config. The
definition of the estimator --- code. The temptation, once you have a config file,
is to make everything configurable, and you end up with a 200-line YAML file harder
to read than the code it configures. The config here has seven settings and covers
everything a reader of the paper would plausibly want to change.

## Keeping the config with the results

One habit worth adopting: since `config.yaml` determines what the numbers are, it is
part of the manuscript's provenance. @sec-manuscript lists it as an input to the
manuscript rule for that reason, and so does the real project these examples are
modeled on. Commit it, and the paper you compiled last March can be rebuilt from
the tag.

## Try it

1. Run with `--config iterations=20`, then again with `--config iterations=200`.
   What does Snakemake rerun, and why?
2. Add `1000` to the `n` list in `config.yaml` and run `snakemake -n`.
3. Delete the `wildcard_constraints` block and ask for
   `rds/sim/n50_rho0_na_ive.rds`. What happens, and what did `{method}` match?
