# Configuration {#sec-config}

```{r}
#| include: false
source("_common.R")
```

The Snakefile in @sec-wildcards has the parameter grid written into it, and `sim.R`
has 200 iterations and the seed written into it. Both are the wrong places. You will
want 20 iterations while you are debugging and 1000 for the final results, and a
coauthor will want to add a sample size without reading your Python.

Move to `examples/03-config`, which puts all of it in one file.

## config.yaml

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/config.yaml")
```

YAML is about as simple as it looks: `key: value`, one per line, lists in square
brackets, `#` for comments. Load it at the top of the Snakefile:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^configfile:", to = "^configfile:",
           filename = "Snakefile (excerpt)")
```

and everything in it is available as `config`, a dictionary --- `config["iterations"]`
is `200` and `config["n"]` is the list `[50, 100, 400]`. Use `config.get("key",
default)` when a setting is optional.

## params

Values from the config reach a shell command through `params`:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^rule sim:", to = '^        "--out',
           filename = "Snakefile (excerpt)")
```

You could write `config["seed"]` straight into the command with a Python f-string
and it would work. Declaring it in `params` is better for one specific reason:
Snakemake records the params it used alongside each output, so changing one marks
the affected files stale, exactly as changing an input file does. Values smuggled
into the command another way are invisible to that machinery.

## params or filename? {#sec-params-or-filename}

Notice that `seed` and `level` are params but `iterations` is not --- it is a
wildcard, and it names a directory. That is the most consequential decision in
this chapter, and it is worth being deliberate about.

Every setting that affects a result has to be visible to Snakemake somehow. There
are exactly two places to put it, and the choice is not about taste:

**In the output path** --- when different values are different results, both of
which you might want to keep. Running 20 iterations and running 1000 does not make
one of them wrong; they are a quick check and a final answer, and there is no
reason they should not sit on disk together.

**In `params:`** --- when there is only one right value, so changing it means the
existing file was simply *wrong* and should be replaced. The seed is like this. You
do not want `seed-1234` and `seed-9999` versions of everything; you want the
results, from the seed the paper reports.

**In neither** is a bug. Snakemake cannot see the change, nothing reruns, and your
results are quietly stale --- the failure mode of @sec-debugging.

So `iterations` goes in the path:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^## The iteration count",
           to = "^ITER = ", filename = "Snakefile (excerpt)")
```

and the smoke run and the real run coexist:

```{r}
#| echo: false
#| output: asis
transcript("03-coexist.txt")
```

Eighteen results at 20 iterations, eighteen at 200, and going back to the config's
default rebuilds only the figure --- the 200-iteration simulations were still there
and still valid. Had `iterations` been a `params` value, the second run would have
overwritten the first, and switching back would have meant recomputing all
eighteen.

`ITER = config["iterations"]` is what selects which of them the figure is built
from, so the config still controls the answer; it just no longer destroys the
alternative. @sec-scaling shows the same pattern at the scale where it matters,
where the recomputation being avoided is measured in days.

The grid comes from the config too, so `expand()` reads:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "        rds = expand",
           to = "                     method=config", filename = "Snakefile (excerpt)")
```

Adding a fourth sample size is now a two-character edit to `config.yaml`.

## wildcard_constraints

@sec-wildcards warned that a wildcard matches anything. Constraints fix that:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^wildcard_constraints:",
           to = '^    method = ', filename = "Snakefile (excerpt)")
```

`r"\d+"` is a regular expression matching one or more digits; the `r` prefix marks a
raw string, so the backslash reaches the regex intact. Declared at the top level
they apply to every rule, and they can also be attached to a single rule, as
@sec-manuscript does.

Two reasons to bother:

**Ambiguity.** With unconstrained wildcards there is more than one way to match
`n{n}_rho{rho}_{method}.rds` against a name containing extra underscores, and
Snakemake resolves it by its rules rather than yours.

**Failing early.** A typo like `n50_rho0_naive.RDS` now fails immediately with "no
rule to produce" instead of launching a job that dies twenty minutes later inside R.

## Overriding from the command line

Any config value can be replaced for a single run with `--config`:

```bash
snakemake --cores 4 --config iterations=20
```

which is the fastest way to check that a pipeline is wired up before committing to
the real thing:

```{r}
#| echo: false
#| output: asis
transcript("03-smoke-vs-full.txt")
```

Five seconds against eleven, for a grid that would take considerably longer at the
iteration counts of a real paper.

::: {.callout-warning}
## `--config` does not always rebuild what you expect

Which of the cases in @sec-params-or-filename a setting falls into decides what
overriding it does, and they behave quite differently:

- **`--config iterations=20`** asks for *different files*. Eighteen new simulations
  run; the old ones are left alone.
- **`--config seed=99`** changes a `params` value. The eighteen existing files are
  marked stale and rebuilt in place.
- **Dropping a sample size from the grid** changes only the list inside `expand()`.
  No simulation reruns --- the remaining results are still perfectly valid --- but
  the figure does, with the reason `Set of input files has changed since last
  execution`, because it is now summarizing twelve files instead of eighteen.

Note that `--config` wants one `key=value` per setting and will not take a list on
the command line (`--config n=[50,100]` is a parse error on 9.13). Change the grid
in `config.yaml`, or point `--configfile` at a second YAML file whose values
override the first.

When you want to force the issue, `snakemake --forcerun sim` reruns every job of
that rule regardless. See @sec-debugging.
:::

## Where to draw the line

A useful test: **if changing it changes the results, it belongs in the config; if
changing it changes what the code does, it belongs in the code.**

Iteration counts, seeds, grids, confidence levels, file locations --- config. The
definition of the estimator --- code. The temptation, once you have a config file,
is to make everything configurable, and you end up with a 200-line YAML file harder
to read than the code it configures. The config here has seven settings and covers
everything a reader of the paper would plausibly want to change.

## Keeping the config with the results

One habit worth adopting: since `config.yaml` determines what the numbers are, it is
part of the manuscript's provenance. @sec-manuscript lists it as an input to the
manuscript rule for that reason, and so does the real project these examples are
modeled on. Commit it, and the paper you compiled last March can be rebuilt from
the tag.

## Try it

1. Run with `--config iterations=20`, then again with `--config iterations=200`.
   What does Snakemake rerun, and why?
2. Add `1000` to the `n` list in `config.yaml` and run `snakemake -n`.
3. Delete the `wildcard_constraints` block and ask for
   `rds/sim/n50_rho0_na_ive.rds`. What happens, and what did `{method}` match?
