# Configuration {#sec-config}

```{r}
#| include: false
source("_common.R")
```

The Snakefile in @sec-wildcards has the parameter grid written into it, and `sim.R`
has 200 iterations and the seed written into it. Both are the wrong places. You will
want 20 iterations while you are debugging and 1000 for the final results, and a
coauthor will want to add a sample size without reading your Python.

Move to `examples/03-config`, which puts all of it in one file.

## config.yaml

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/config.yaml")
```

YAML is about as simple as it looks: `key: value`, one per line, lists in square
brackets, `#` for comments. Load it at the top of the Snakefile:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^configfile:", to = "^configfile:",
           filename = "Snakefile (excerpt)")
```

and everything in it is available as `config`, a dictionary --- `config["iterations"]`
is `200` and `config["n"]` is the list `[50, 100, 400]`. Use `config.get("key",
default)` when a setting is optional.

## Using config values in a rule

Pull the settings out into plain Python variables at the top of the Snakefile, and
use them in the shell command like any other placeholder:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^ITER  = ", to = "^LEVEL = ",
           filename = "Snakefile (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^rule sim:", to = '^        "--out',
           filename = "Snakefile (excerpt)")
```

`{SEED}` and `{LEVEL}` are filled in from those variables, exactly as
`{wildcards.n}` is filled in from the wildcard. Anything defined at the top of the
Snakefile is available this way.

## Runs that don't overwrite each other {#sec-iterations-path}

One setting is treated differently, and it is the most useful idea in this chapter.
`iterations` is not passed to the script and forgotten --- it names the directory
the results go in:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^## The iteration count",
           to = "^ITER", filename = "Snakefile (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^    output:", to = "^        \"rds/",
           filename = "Snakefile (excerpt)")
```

The reason is that 20 iterations and 1000 iterations are not a right answer and a
wrong one. They are a quick check and a final result, and there is no reason they
should not sit on disk at the same time. Because the count is part of the path,
they do:

```{r}
#| echo: false
#| output: asis
transcript("03-coexist.txt")
```

Eighteen results at 20 iterations, eighteen at 200, and going back to the config's
default rebuilds only the figure --- the 200-iteration simulations were still there
and still valid. Had the count merely been handed to the script, the smoke run
would have overwritten the real one, and switching back would have meant
recomputing all eighteen.

`ITER = config["iterations"]` is what selects which set the figure is built from,
so the config still controls the answer; it just no longer destroys the
alternative. @sec-scaling shows the same pattern at the scale where it matters,
where the recomputation being avoided is measured in days.

The grid comes from the config too, so `expand()` reads:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "        rds = expand",
           to = "                     method=config", filename = "Snakefile (excerpt)")
```

Adding a fourth sample size is now a two-character edit to `config.yaml`.

## wildcard_constraints

@sec-wildcards warned that a wildcard matches anything. Constraints fix that:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^wildcard_constraints:",
           to = '^    method = ', filename = "Snakefile (excerpt)")
```

`r"\d+"` is a regular expression matching one or more digits; the `r` prefix marks a
raw string, so the backslash reaches the regex intact. Declared at the top level
they apply to every rule, and they can also be attached to a single rule, as
@sec-manuscript does.

Two reasons to bother:

**Ambiguity.** With unconstrained wildcards there is more than one way to match
`n{n}_rho{rho}_{method}.rds` against a name containing extra underscores, and
Snakemake resolves it by its rules rather than yours.

**Failing early.** A typo like `n50_rho0_naive.RDS` now fails immediately with "no
rule to produce" instead of launching a job that dies twenty minutes later inside R.

## Overriding from the command line

Any config value can be replaced for a single run with `--config`:

```bash
snakemake --cores 4 --config iterations=20
```

which is the fastest way to check that a pipeline is wired up before committing to
the real thing:

```{r}
#| echo: false
#| output: asis
transcript("03-smoke-vs-full.txt")
```

Five seconds against eleven, for a grid that would take considerably longer at the
iteration counts of a real paper.

::: {.callout-warning}
## `--config` does not always rebuild what you expect

What an override actually rebuilds depends on how the setting is used, and the two
cases here behave differently:

- **`--config iterations=20`** asks for *different files*. Eighteen new simulations
  run; the old ones are left alone.
- **Dropping a sample size from the grid** changes only the list inside `expand()`.
  No simulation reruns --- the remaining results are still perfectly valid --- but
  the figure does, with the reason `Set of input files has changed since last
  execution`, because it is now summarizing twelve files instead of eighteen.

Note that `--config` wants one `key=value` per setting and will not take a list on
the command line (`--config n=[50,100]` is a parse error on 9.13). Change the grid
in `config.yaml`, or point `--configfile` at a second YAML file whose values
override the first.

When you want to force the issue, `snakemake --forcerun sim` reruns every job of
that rule regardless. See @sec-debugging.
:::

## Where to draw the line

A useful test: **if changing it changes the results, it belongs in the config; if
changing it changes what the code does, it belongs in the code.**

Iteration counts, seeds, grids, confidence levels, file locations --- config. The
definition of the estimator --- code. The temptation, once you have a config file,
is to make everything configurable, and you end up with a 200-line YAML file harder
to read than the code it configures. The config here has seven settings and covers
everything a reader of the paper would plausibly want to change.

## Keeping the config with the results

One habit worth adopting: since `config.yaml` determines what the numbers are, it is
part of the manuscript's provenance. @sec-manuscript lists it as an input to the
manuscript rule for that reason, and so does the real project these examples are
modeled on. Commit it, and the paper you compiled last March can be rebuilt from
the tag.

## Try it

1. Run with `--config iterations=20`, then again with `--config iterations=200`.
   What does Snakemake rerun, and why?
2. Add `1000` to the `n` list in `config.yaml` and run `snakemake -n`.
3. Delete the `wildcard_constraints` block and ask for
   `rds/sim/n50_rho0_na_ive.rds`. What happens, and what did `{method}` match?
