# Configuration {#sec-config}

```{r}
#| include: false
source("_common.R")
```

`sim.R` in @sec-wildcards has 200 iterations written into it, which is the wrong
place: you will want 20 while you are debugging and 1000 for the final results, and
changing that should not mean editing the simulation code.

Move to `examples/03-config`, which puts all of it in one file.

## config.yaml

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/config.yaml")
```

YAML is about as simple as it looks: `key: value`, one per line, lists in square
brackets, `#` for comments. Load it at the top of the Snakefile:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^configfile:", to = "^configfile:",
           filename = "Snakefile (excerpt)")
```

and everything in it is available as `config`, a dictionary: `config["iterations"]`
is `200`. Use `config.get("key", default)` when a setting is optional.

Notice how little is in there. A config is for settings that apply to the project
as a whole --- `iterations` means the same thing to every rule that computes
anything.

## Runs that don't overwrite each other {#sec-iterations-path}

This one setting is treated differently, and it is the most useful idea in this chapter.
`iterations` is not passed to the script and forgotten --- it names the directory
the results go in:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^## The iteration count",
           to = "^ITER", filename = "Snakefile (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^rule sim:", to = '^        "--out',
           filename = "Snakefile (excerpt)")
```

`{ITER}` never appears in that rule: the count arrives as the wildcard
`{wildcards.iterations}`, read back out of the filename that was asked for by `rule figure`, just
like `{wildcards.n}`.

The figure rule collects the set `ITER` points at:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "        rds = expand",
           to = "                     method = ", filename = "Snakefile (excerpt)")
```

## wildcard_constraints

@sec-wildcards warned that a wildcard matches anything. Constraints fix that:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^wildcard_constraints:",
           to = '^    method = ', filename = "Snakefile (excerpt)")
```

`r"\d+"` is a regular expression matching one or more digits; the `r` prefix marks a
raw string, so the backslash reaches the regex intact. Declared at the top level
they apply to every rule, and they can also be attached to a single rule, as
@sec-manuscript does.

Two reasons to bother:

**Ambiguity.** With unconstrained wildcards there is more than one way to match
`n{n}_rho{rho}_{method}.rds` against a name containing extra underscores, and
Snakemake resolves it by its rules rather than yours.

**Failing early.** Without the `\d+` on `{n}`, asking for
`rds/200/sim/nfifty_rho0_naive.rds` matches quite happily ---

```{.default}
wildcards: iterations=200, n=fifty, rho=0, method=naive
```

--- and Snakemake goes off and runs `sim.R --n fifty`, which dies inside R however
long the job takes to get that far. With the constraint, nothing can produce that
file and you are told so before anything starts.

A misspelled *extension* is a different matter: `n50_rho0_naive.RDS` fails whether
or not you have constraints, because the pattern ends in a literal `.rds` and
`.RDS` is not that. Constraints police the wildcards, not the text around them.

## Overriding from the command line

Any config value can be replaced for a single run with `--config`:

```bash
snakemake --cores 4 --config iterations=20
```

which is the fastest way to check that a pipeline is wired up before committing to
the real thing:

```{r}
#| echo: false
#| output: asis
transcript("03-smoke-vs-full.txt")
```

Five seconds against eleven, for a grid that would take considerably longer at the
iteration counts of a real paper.

::: {.callout-warning}
## `--config` does not always rebuild what you expect

What an override actually rebuilds depends on how the setting is used, and the two
cases here behave differently:

- **`--config iterations=20`** asks for *different files*. Eighteen new simulations
  run; the old ones are left alone.
- **Dropping a sample size from a rule's `expand()`** is a Snakefile edit rather
  than a config one, and it reruns no simulations at all --- the remaining results
  are still perfectly valid. The figure does rerun, with the reason `Set of input
  files has changed since last execution`, because it is now summarizing twelve
  files instead of eighteen.

Note that `--config` wants one `key=value` per setting and will not take a list on
the command line (`--config n=[50,100]` is a parse error on 9.13). Change the grid
in `config.yaml`, or point `--configfile` at a second YAML file whose values
override the first.

When you want to force the issue, `snakemake --forcerun sim` reruns every job of
that rule regardless. See @sec-debugging.
:::

## Where to draw the line

A useful test: **if changing it changes the results, it belongs in the config; if
changing it changes what the code does, it belongs in the code.**

Iteration counts and file locations --- config. The definition of the estimator ---
code. The temptation, once you have a config file, is to make everything
configurable, and you end up with a 200-line YAML file harder to read than the code
it configures. The config here has one setting, and it is the one that genuinely
applies to everything.

## Keeping the config with the results

One habit worth adopting: since `config.yaml` determines what the numbers are, it is
part of the manuscript's provenance. @sec-manuscript lists it as an input to the
manuscript rule for that reason, and so does the real project these examples are
modeled on. Commit it, and the paper you compiled last March can be rebuilt from
the tag.

## Try it

1. Run with `--config iterations=20`, then again with `--config iterations=200`.
   What does Snakemake rerun, and why?
2. Add `1000` to the `n` list in `config.yaml` and run `snakemake -n`.
3. Delete the `wildcard_constraints` block and ask for
   `rds/sim/n50_rho0_na_ive.rds`. What happens, and what did `{method}` match?
