# One rule, many jobs {#sec-wildcards}

```{r}
#| include: false
source("_common.R")
```

Look again at the Snakefile from @sec-first-snakefile and notice how much of it
is copied and pasted:

```{r}
#| echo: false
#| output: asis
embed_file("examples/01-first-rules/Snakefile", from = "^rule sim_naive",
           to = '^        "Rscript sim.R split', filename = "Snakefile (excerpt)")
```

Two rules, identical but for the word `naive` versus `split`. Now imagine the real
version of this simulation: three sample sizes, three correlation levels, two
methods. Eighteen rules, all the same, and one typo away from a figure that
silently mixes up two scenarios.

Wildcards fix this. Move to `examples/02-wildcards`.

## A wildcard is a hole in a filename

```{r}
#| echo: false
#| output: asis
embed_file("examples/02-wildcards/Snakefile", from = "^rule sim:",
           to = '^        "--out \\{output\\}"', filename = "Snakefile (excerpt)")
```

The output is no longer a filename; it is a *pattern*. The parts in braces are
wildcards, and they work backwards: when something asks for
`rds/sim/n400_rho80_split.rds`, Snakemake matches that name against the
pattern, concludes that `n` is `400`, `rho` is `80` and `method` is `split`, and
substitutes those values into the command wherever `{wildcards.…}` appears.

This is the mental adjustment that makes Snakemake click. You do not tell it to run
the simulation eighteen times. You tell it what a result file is *called* as a
function of its parameters, and then you ask for the files you want.

::: {.callout-important}
## Wildcards match anything, including nonsense

A wildcard is a regular expression, `.+` by default. Nothing in the rule above says
`method` has to be `naive` or `split`, so asking for
`rds/sim/n50_rho0_bayes.rds` produces a perfectly valid job that runs `sim.R
--method bayes` and fails inside R. @sec-config shows how to constrain wildcards,
and @sec-debugging shows what that failure looks like.
:::

## Naming files so this works

Wildcards make your file-naming scheme part of the pipeline's logic, which is worth
two minutes of thought up front:

- **Put every parameter that changes into the name.** If two runs differ in
  anything, their outputs must differ in name, or one will overwrite the other and
  Snakemake will believe both are up to date.
- **Use a separator that cannot appear in a value.** These examples use `_` between
  fields and stick to digits and lowercase letters within a field.

The scheme here is `n{n}_rho{rho}_{method}.rds`, with `rho` as a percentage so that
it stays an integer --- `rho50` rather than `rho0.5`, because a decimal point in a
filename is an invitation to trouble.

## expand()

Something still has to say which eighteen files we want. That is `expand()`:

```{r}
#| echo: false
#| output: asis
embed_file("examples/02-wildcards/Snakefile", from = "^N = ", to = "^METHODS",
           filename = "Snakefile (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/02-wildcards/Snakefile", from = "^rule figure:",
           filename = "Snakefile (excerpt)")
```

`expand()` takes a pattern and some lists and returns every combination of them, so

```python
expand("rds/sim/n{n}_rho{rho}_{method}.rds", n=N, rho=RHO, method=METHODS)
```

returns a list of 3 × 3 × 2 = 18 filenames. Those are the inputs of `rule figure`,
which is what causes all eighteen simulations to run.

::: {.callout-note}
## The Python you need

`N = [50, 100, 400]` is a list, the same idea as `c(50, 100, 400)` in R. That is
essentially the entire Python vocabulary required to use Snakemake: lists, strings
in quotes, and the occasional dictionary in @sec-config.
:::

The `expand()` sits inside the rule that needs it, which is where a grid usually
belongs: another rule is free to sweep a different one, and a supplementary figure
covering only the large sample sizes would simply expand over fewer. When several
rules genuinely share a grid, define it once above them and use the variable in
each.

The script on the other end receives all eighteen filenames as arguments and
stacks them:

```{r}
#| echo: false
#| output: asis
embed_file("examples/02-wildcards/figure.R", from = "^args <- commandArgs",
           to = "^res <- do.call", filename = "figure.R (excerpt)")
```

Note that `expand()` fills wildcards in, and `{wildcards.n}` reads them back out.
These are opposite operations, and it is worth being clear about which one you are
doing: `expand()` appears in the rule that *consumes* many files, and
`{wildcards.…}` in the rule that *produces* one.

## Running it

```{r}
#| echo: false
#| output: asis
transcript("02-job-stats.txt")
```

Twenty jobs from three rules. On four cores:

```{r}
#| echo: false
#| output: asis
transcript("02-run-tail.txt")
```

The DAG has changed shape: one rule now stands for eighteen independent jobs, all of
which must finish before the figure runs.

```{mermaid}
%%| fig-align: center
flowchart LR
  s1["sim<br/>n: 50, rho: 0<br/>method: naive"] --> figure(figure)
  s2["sim<br/>n: 50, rho: 0<br/>method: split"] --> figure
  s3["(15 more sim jobs)"] --> figure
  s4["sim<br/>n: 400, rho: 80<br/>method: split"] --> figure
  figure --> allrule(all)
```

That is the shape `snakemake --dag` reports, abbreviated in the middle: one node
per *job*, each `sim` node carrying the wildcard values it was matched with.
`snakemake --rulegraph` collapses all eighteen back into a single `sim` box, which
is the view you want once a workflow has more than a few rules (@sec-debugging).

## Building one cell of the grid

Because the output is a pattern, you can ask for any file it can match, whether or
not anything else wants it:

```{r}
#| echo: false
#| output: asis
transcript("02-one-target.txt")
```

This is how you develop. Rather than running the whole grid to find out whether
your new method works, ask for one small scenario, look at it, and only then let
the rest go.

This is also extremely powerful, because now if you want a certain simulation... you just ahve to ask for it rather than edit an R file and run it.

## Try it

1. Add `1000` to `N` and run `snakemake -n`. How many jobs? Which ones are new?
2. Ask for `rds/sim/n77_rho33_naive.rds`. It works --- explain why.
3. Ask for `rds/sim/n50_rho0_bayes.rds` and watch it fail. Where does the error
   come from, and how would you have preferred to find out?
