# Configuration {#sec-config}

```{r}
#| include: false
source("_common.R")
```

`sim.R` in @sec-wildcards has 200 iterations written into it, which is the wrong
place: you will want 20 while you are debugging and 1000 for the final results, and
changing that should not mean editing the simulation code.

Move to `examples/03-config`, which puts shared parameters into a single file.

## config.yaml

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/config.yaml")
```

YAML is about as simple as it looks: `key: value`, one per line, lists in square
brackets, `#` for comments. Load it at the top of the Snakefile:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^configfile:", to = "^configfile:",
           filename = "Snakefile (excerpt)")
```

and everything in it is available as `config`, a dictionary: `config["iterations"]`
is `200`. Use `config.get("key", default)` when a setting is optional.

Notice how little is in there. A config is for settings that apply to the project
as a whole --- `iterations` means the same thing to every simulation.

## Runs that don't overwrite each other {#sec-iterations-path}

This one setting is treated differently, and it is the most useful idea in this chapter.
`iterations` is not passed to the script and forgotten --- it names the directory
the results go in:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^## The iteration count",
           to = "^ITER", filename = "Snakefile (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "^rule sim:", to = '^        "--out',
           filename = "Snakefile (excerpt)")
```

`{ITER}` never appears in that rule: the count arrives as the wildcard
`{wildcards.iterations}`, read back out of the filename that was asked for by `rule figure`, just
like `{wildcards.n}`.

The figure rule collects the set `ITER` points at:

```{r}
#| echo: false
#| output: asis
embed_file("examples/03-config/Snakefile", from = "        rds = expand",
           to = "                     method = ", filename = "Snakefile (excerpt)")
```

## Overriding from the command line

Any config value can be replaced for a single run with `--config`:

```bash
snakemake --cores 4 --config iterations=20
```

which is the fastest way to check that a pipeline is wired up before committing to
the real thing:

```{r}
#| echo: false
#| output: asis
transcript("03-smoke-vs-full.txt")
```

Five seconds against eleven, for a grid that would take considerably longer at the
iteration counts of a real paper.

## Try it

1. Run with `--config iterations=20`, then again with `--config iterations=200`.
   What does Snakemake rerun, and why?
3. Delete the `wildcard_constraints` block and ask for
   `rds/sim/n50_rho0_na_ive.rds`. What happens, and what did `{method}` match?
