# Parallelism, logs and benchmarks {#sec-parallel}

```{r}
#| include: false
source("_common.R")
```

Simulation studies are the easiest parallel problem there is: a thousand independent
replications, no communication between them. The reason to mention this in a
Snakemake tutorial is that once your work is expressed as a DAG, you get the
parallelism for free --- Snakemake already knows which jobs do not depend on each
other.

## --cores

`--cores N` is a promise about the whole machine, not a setting per rule: Snakemake
will keep up to `N` cores busy, starting whatever jobs are runnable.

```bash
snakemake --cores 4
```

With eighteen independent simulations and four cores, four run, and as each
finishes another starts, until all eighteen are done and the figure can go. There is
nothing to configure --- the ordering falls out of the graph.

`--cores all` uses every core you have. On a laptop you are also using for other
things, leave one or two free; on a shared server, be a good citizen and pick a
number.

::: {.callout-note}
## Two kinds of parallel

`--cores` controls how many *jobs* run at once, and that is usually the only kind
of parallelism you need for a simulation study. It is unrelated to parallelism
*inside* a job (`mclapply`, `future`, a multithreaded BLAS). If a rule does use
threads internally, tell Snakemake with `threads:` so its accounting stays honest:

```python
rule sim:
    threads: 4
    shell: "Rscript sim.R --cores {threads} ..."
```

Then `--cores 8` runs two such jobs at a time rather than eight. Getting this wrong
in the other direction --- eight jobs each spawning eight threads on an eight-core
machine --- is the single most common way to make a pipeline slower by parallelizing
it.
:::

## Logs

Once eighteen jobs are interleaving their output, the terminal becomes unreadable,
and the run you actually want to look at scrolled past an hour ago. The `log:`
directive gives each job its own file:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule sim:",
           to = '^        "--out \\{output\\} > \\{log\\} 2>&1"',
           filename = "Snakefile (excerpt)")
```

Two details. The log file is *declared* by `log:` but nothing writes to it
automatically --- the redirection `> {log} 2>&1` at the end of the shell command is
what does it (`2>&1` catches errors as well as normal output). And the log path
carries the same wildcards as the output, so each job gets its own file rather than
eighteen jobs overwriting one.

Anything the script prints lands there:

```{r}
#| echo: false
#| output: asis
transcript("04-log.txt")
```

Which is worth designing for. A script that prints its parameters, its sample size
after exclusions, and a summary of what it computed gives you a log directory that
answers most questions without rerunning anything. @sec-analysis leans on this
heavily.

Logs also survive failure. When a job dies, Snakemake tells you which log to read,
and the R error message is sitting in it --- as opposed to being lost in the
scrollback of a parallel run.

::: {.callout-important}
## Snakemake will not delete a log file

Outputs of a failed job are removed so that a half-written file is never mistaken
for a real one. Logs are deliberately kept, because they are the evidence. This also
means a log from a successful run sticks around until the job runs again.
:::

## Benchmarks

`benchmark:` records how long a job took and how much memory it used:

```python
rule sim:
    benchmark: "benchmarks/sim/n{n}_rho{rho}_{method}.tsv"
```

```{r}
#| echo: false
#| output: asis
transcript("04-benchmark.txt")
```

The first column is seconds. This is how you find out that one corner of your
parameter grid takes forty times as long as the rest, which is worth knowing before
you scale the whole grid up by a factor of ten.

The memory columns come from `/proc` and are populated on Linux; on macOS they read
zero, as above. Timing works everywhere.

## Interrupting and resuming

Snakemake is designed to be stopped. `Ctrl-C` lets running jobs finish and shuts
down cleanly; anything already built stays built, and the next run picks up where
this one left off. That property --- being able to kill a twelve-hour simulation
after four hours and lose only the jobs that were mid-flight --- is worth more in
practice than the parallelism.

If you kill it *hard* (`kill -9`, a closed laptop, a cluster timeout), a job may
have been interrupted with its output file half-written. Snakemake notices and
refuses to proceed until you say what to do:

```bash
snakemake --cores 4 --rerun-incomplete
```

which discards the suspect files and rebuilds them.

## Try it

1. In `examples/03-config`, run with `--cores 1`, then `--cores 4`, and compare.
   Why is the speedup less than four?
2. Delete one simulation's `.rds` file and run again. How many jobs run?
3. Add a `log:` directive to the `figure` rule in `examples/03-config` and check
   that `ggsave`'s output ends up in it.
