# Figures and tables {#sec-figures}

```{r}
#| include: false
source("_common.R")
```

Everything so far has produced `.rds` files, which nobody reads. This chapter turns
them into the two things a paper is actually made of. Move to
`examples/04-manuscript`, which is the complete version of the simulation study.

## The aggregating rule

A figure rule looks like any other rule; what distinguishes it is that its input is
the whole grid rather than one cell:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule figure1:", to = '^        "\\{input.rds\\} > \\{log\\} 2>&1"',
           filename = "Snakefile (excerpt)")
```

The `expand()` sits in the rule that needs it. `rule table1` next to it writes out
the same call, because it happens to summarize the same grid --- but it is free not
to, and a supplementary figure covering only the large sample sizes would simply
sweep less.

The R script receives all eighteen filenames as trailing arguments and stacks them:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/figure.R", from = "^parsed <- parse_args",
           to = "^res <- do.call", filename = "code/figure.R (excerpt)")
```

`positional_arguments = TRUE` tells `optparse` to accept bare arguments alongside
the named flags, so the script can be called as

```bash
Rscript code/figure.R --out out/figure1.pdf rds/200/sim/*.rds
```

An alternative worth knowing about: instead of passing filenames, the script could
glob `rds/200/sim/` itself. Do not do this. A glob picks up whatever happens to be
lying in the directory, including results from a parameter grid you abandoned last
month, and it hides the dependency from Snakemake, which will then happily build
your figure without running the simulations.

## Figures

Nothing about the plotting code is Snakemake-specific --- it is the `ggsave()` call
you would write anyway:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/figure.R", from = "^p <- ggplot",
           to = "^cat\\(\"wrote", filename = "code/figure.R (excerpt)")
```

Two conventions, though, that matter more inside a pipeline than outside one:

**Write vector graphics to the size they will appear.** `ggsave(..., width = 7,
height = 3)` in inches, matching the text width of the document, means the figure is
never rescaled by LaTeX and the fonts in it come out the size you chose. Resizing a
figure by 60% in `\includegraphics` is how you end up with axis labels nobody can
read.

**Name the methods once, at the top of the script.** The figure and the table both
turn `naive` into something a reader sees:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/figure.R", from = "^method_labels",
           to = "^colors", filename = "code/figure.R (excerpt)")
```

so that the spelling and the color are decided in one visible place rather than
buried in a `ggplot` call. In a project with more than a couple of figures it is
worth going further and putting these in a file the scripts share; here, where each
script is short enough to read at a glance, keeping them self-contained is simpler.

::: {.callout-note}
## Shared code belongs in a file of its own

That goes double for functions. `sim.R` defines `gen_data()` and
`select_forward()` inline because nothing else in this project needs them, and the
script stays readable top to bottom. The moment a second script wants one of them
--- a supplementary simulation, a figure that regenerates a single dataset to
illustrate the method --- copying it across is how two versions of your estimator
begin to disagree.

At that point give them a file, say `code/functions.R`, and `source()` it. There is
one Snakemake-specific thing to remember when you do: list it as an input on every
rule whose script sources it.

```python
rule sim:
    input:
        script = "code/sim.R",
        functions = "code/functions.R"
```

Without that line the dependency is invisible, and editing a function leaves every
result that used it silently stale --- the one failure mode Snakemake cannot warn
you about (@sec-debugging).

When the functions are the contribution rather than the scaffolding, the natural
home is an R package rather than a loose file, which is where @sec-scaling picks
the story up.
:::

## Tables

Tables are the same idea with a different output format: the script writes a LaTeX
fragment that the manuscript `\input`s.

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/table.R", from = "^## escape = FALSE",
           to = "^  add_header_above", filename = "code/table.R (excerpt)")
```

which produces:

```{r}
#| echo: false
#| output: asis
embed_file("transcripts/table1.tex", lang = "latex",
           filename = "out/table1.tex")
```

Points worth noting:

- The file contains a **fragment**, not a document --- no `\documentclass`, no
  `\begin{document}`. It is meant to be `\input` into something else.
- The **caption and label live with the numbers**, not in the manuscript. That way
  `\ref{tab:coverage}` resolves and the caption cannot describe a table that the
  code no longer produces.
- `escape = FALSE` lets `$n$` through as math. The cost is that everything else in
  the table must then be valid LaTeX --- a `%` or `_` in a variable name will break
  the build, which is a reasonable trade when you control the contents.

Writing `.tex` rather than pasting numbers into the manuscript by hand is the entire
argument of this book, applied to tables. A number in your paper should be
traceable to the code that computed it, and the only way to guarantee that is for
no human hand to touch it in between.

## rule all as the manifest

With figures and tables in place, `rule all` becomes the list of everything the
paper is made of:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule all:", to = '^        "build/submission.pdf"',
           filename = "Snakefile (excerpt)")
```

Here it names only the two PDFs, because everything else is reachable through them.
In a larger project --- the one in @sec-scaling has seventeen figures and tables ---
it is common to list every product explicitly instead:

```python
rule all:
    input:
        "build/paper.pdf",
        "out/figure1.pdf",
        "out/figure2.pdf",
        ...
```

This is slightly redundant, and it is what you want anyway. It gives you one place
to read off what the project produces, and it means `snakemake` still builds a
figure you have temporarily commented out of the manuscript while you rearrange
the results section.

## Try it

1. Change the color palette in `code/figure.R` and run `snakemake -n`. Does the
   table rebuild? Should it?
2. Add a `figure2` rule showing coverage against `rho` instead of `n`, and add it
   to `rule all`.
3. Delete `out/table1.tex` and run `snakemake --cores 4`. How far back does
   Snakemake go?
