# Figures and tables {#sec-figures}

```{r}
#| include: false
source("_common.R")
```

Everything so far has produced `.rds` files, which nobody reads. This chapter turns
them into the two things a paper is actually made of. Move to
`examples/04-manuscript`, which is the complete version of the simulation study.

## The aggregating rule

A figure rule looks like any other rule; what distinguishes it is that its input is
the whole grid rather than one cell:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^## Everything the manuscript",
           to = "^              n=config", filename = "Snakefile (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule figure1:", to = '^        "\\{input.rds\\} > \\{log\\} 2>&1"',
           filename = "Snakefile (excerpt)")
```

Pulling the `expand()` out into a variable named `SIMS` is worth doing as soon as
more than one rule needs the same list. The figure and the table both summarize the
whole grid, and defining that list once means they cannot drift apart.

The R script receives all eighteen filenames as trailing arguments and stacks them:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/figure.R", from = "^parsed <- parse_args",
           to = "^res <- do.call", filename = "code/figure.R (excerpt)")
```

`positional_arguments = TRUE` tells `optparse` to accept bare arguments alongside
the named flags, so the script can be called as

```bash
Rscript code/figure.R --out out/figure1.pdf --level 0.95 rds/sim/*.rds
```

An alternative worth knowing about: instead of passing filenames, the script could
glob `rds/sim/` itself. Do not do this. A glob picks up whatever happens to be
lying in the directory, including results from a parameter grid you abandoned last
month, and it hides the dependency from Snakemake, which will then happily build
your figure without running the simulations.

## Figures

Nothing about the plotting code is Snakemake-specific --- it is the `ggsave()` call
you would write anyway:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/figure.R", from = "^p <- ggplot",
           to = "^cat\\(\"wrote", filename = "code/figure.R (excerpt)")
```

Two conventions, though, that matter more inside a pipeline than outside one:

**Write vector graphics to the size they will appear.** `ggsave(..., width = 7,
height = 3)` in inches, matching the text width of the document, means the figure is
never rescaled by LaTeX and the fonts in it come out the size you chose. Resizing a
figure by 60% in `\includegraphics` is how you end up with axis labels nobody can
read.

**Keep the appearance in `setup.R`.** Colors and method labels live in one place:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/setup.R", from = "^method_labels",
           to = "^colors", filename = "code/setup.R (excerpt)")
```

so that "naive" is the same color in every figure and is spelled the same way in
the table. This is a small thing that becomes a large thing at revision time.

## Tables

Tables are the same idea with a different output format: the script writes a LaTeX
fragment that the manuscript `\input`s.

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/table.R", from = "^## escape = FALSE",
           to = "^  add_header_above", filename = "code/table.R (excerpt)")
```

which produces:

```{r}
#| echo: false
#| output: asis
embed_file("transcripts/table1.tex", lang = "latex",
           filename = "out/table1.tex")
```

Points worth noting:

- The file contains a **fragment**, not a document --- no `\documentclass`, no
  `\begin{document}`. It is meant to be `\input` into something else.
- The **caption and label live with the numbers**, not in the manuscript. That way
  `\ref{tab:coverage}` resolves and the caption cannot describe a table that the
  code no longer produces.
- `escape = FALSE` lets `$n$` through as math. The cost is that everything else in
  the table must then be valid LaTeX --- a `%` or `_` in a variable name will break
  the build, which is a reasonable trade when you control the contents.

Writing `.tex` rather than pasting numbers into the manuscript by hand is the entire
argument of this book, applied to tables. A number in your paper should be
traceable to the code that computed it, and the only way to guarantee that is for
no human hand to touch it in between.

## rule all as the manifest

With figures and tables in place, `rule all` becomes the list of everything the
paper is made of:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule all:", to = '^        "build/submission.pdf"',
           filename = "Snakefile (excerpt)")
```

Here it names only the two PDFs, because everything else is reachable through them.
In a larger project --- the one in @sec-scaling has seventeen figures and tables ---
it is common to list every product explicitly instead:

```python
rule all:
    input:
        "build/paper.pdf",
        "out/figure1.pdf",
        "out/figure2.pdf",
        ...
```

This is slightly redundant, and it is what you want anyway. It gives you one place
to read off what the project produces, and it means `snakemake` still builds a
figure you have temporarily commented out of the manuscript while you rearrange
the results section.

## Try it

1. Change the color palette in `code/setup.R` and run `snakemake -n`. Does the
   table rebuild? Should it?
2. Add a `figure2` rule showing coverage against `rho` instead of `n`, and add it
   to `rule all`.
3. Delete `out/table1.tex` and run `snakemake --cores 4`. How far back does
   Snakemake go?
