# Figures and tables {#sec-figures}

```{r}
#| include: false
source("_common.R")
```

Everything so far has produced `.rds` files, which nobody reads. This chapter turns
them into the two things a paper is actually made of. Move to
`examples/04-manuscript`, which is the complete version of the simulation study.

## The aggregating rule

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule figure1:", to = '^        "\\{input.rds\\} > \\{log\\} 2>&1"',
           filename = "Snakefile (excerpt)")
```

The `expand()` sits in the rule that needs it. `rule table1` next to it writes out
the same call, because it happens to summarize the same grid --- but it is free not
to, and a supplementary figure covering only the large sample sizes would simply
sweep less.

If multiple figures / tables share the same grid, they can certainly be defined once outside individual rules.

The R script receives all eighteen filenames as trailing arguments and stacks them:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/scripts/figure.R", from = "^parsed <- parse_args",
           to = "^res <- do.call", filename = "scripts/figure.R (excerpt)")
```

`positional_arguments = TRUE` tells `optparse` to accept bare arguments alongside
the named flags, so the script can be called as

```bash
Rscript scripts/figure.R --out out/figure1.pdf rds/200/sim/*.rds
```

## Figures

Nothing about the plotting code is Snakemake-specific --- it is the `ggsave()` call
you would write anyway:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/scripts/figure.R", from = "^p <- ggplot",
           to = "^cat\\(\"wrote", filename = "scripts/figure.R (excerpt)")
```

Two conventions, though, that matter more inside a pipeline than outside one:

**Keep the appearance in one place.** The figure and the table both turn `naive`
into something a reader sees, so the label and the color are decided once, in
`scripts/setup.R`, which the scripts that draw things source:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/scripts/setup.R", from = "^method_labels",
           to = "^colors", filename = "scripts/setup.R (excerpt)")
```

rather than being buried in two separate `ggplot` calls that can drift apart.

::: {.callout-note}
## Declare what a script sources --- and split it so that hurts less

Snakemake has no way to know that `figure.R` reads another file, so `scripts/setup.R`
is listed in the `input:` of the rules whose scripts source it. An undeclared
dependency is the one failure mode Snakemake cannot warn you about: change a color
and the figure that uses it stays quietly stale (@sec-debugging). Declare it
wherever changing it could change an output.

Which is an argument for keeping such files narrow. `setup.R` here holds *only*
presentation --- labels, colors, the plotting packages --- and the simulation rules
do not use `setup.R`. The simulation functions live somewhere else entirely, one
per file:

```{.default filename="examples/04-manuscript"}
R/gen_data.R          sourced by scripts/sim.R      -> an input to rule sim
R/select_forward.R
R/one_rep.R
scripts/setup.R          sourced by figure.R and table.R
```

so that the two kinds of change cost what they should. Editing a hex code plans
four jobs --- the figure, the table and the two PDFs --- and leaves the grid alone.
Editing `R/one_rep.R` plans all twenty-two, because the results really are wrong
now. Had both lived in one `setup.R`, the color change would have rerun eighteen unnecessary
simulations, since Snakemake cannot tell which half of a file you touched.

The `R/` directory is the layout an R package uses.
[IowaBiostat/project-workflow](https://github.com/IowaBiostat/project-workflow)
is a talk and a small example project that goes into this further and is worth giving a read.
:::

## Try it

1. Change the color palette in `scripts/setup.R` and run `snakemake -n`; then do the
   same to `R/gen_data.R`. Count the jobs each time, and explain the difference.
2. Add a `figure2` rule showing coverage against `rho` instead of `n`, and add it
   to `rule all`.
3. Delete `out/table1.tex` and run `snakemake --cores 4`. How far back does
   Snakemake go?
