# Figures and tables {#sec-figures}

```{r}
#| include: false
source("_common.R")
```

Everything so far has produced `.rds` files, which nobody reads. This chapter turns
them into the two things a paper is actually made of. Move to
`examples/04-manuscript`, which is the complete version of the simulation study.

## The aggregating rule

A figure rule looks like any other rule; what distinguishes it is that its input is
the whole grid rather than one cell:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule figure1:", to = '^        "\\{input.rds\\} > \\{log\\} 2>&1"',
           filename = "Snakefile (excerpt)")
```

The `expand()` sits in the rule that needs it. `rule table1` next to it writes out
the same call, because it happens to summarize the same grid --- but it is free not
to, and a supplementary figure covering only the large sample sizes would simply
sweep less.

If multiple figures / tables

The R script receives all eighteen filenames as trailing arguments and stacks them:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/figure.R", from = "^parsed <- parse_args",
           to = "^res <- do.call", filename = "code/figure.R (excerpt)")
```

`positional_arguments = TRUE` tells `optparse` to accept bare arguments alongside
the named flags, so the script can be called as

```bash
Rscript code/figure.R --out out/figure1.pdf rds/200/sim/*.rds
```

An alternative worth knowing about: instead of passing filenames, the script could
glob `rds/200/sim/` itself. Do not do this. A glob picks up whatever happens to be
lying in the directory, including results from a parameter grid you abandoned last
month, and it hides the dependency from Snakemake, which will then happily build
your figure without running the simulations.

## Figures

Nothing about the plotting code is Snakemake-specific --- it is the `ggsave()` call
you would write anyway:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/figure.R", from = "^p <- ggplot",
           to = "^cat\\(\"wrote", filename = "code/figure.R (excerpt)")
```

Two conventions, though, that matter more inside a pipeline than outside one:

**Write vector graphics to the size they will appear.** `ggsave(..., width = 7,
height = 3)` in inches, matching the text width of the document, means the figure is
never rescaled by LaTeX and the fonts in it come out the size you chose. Resizing a
figure by 60% in `\includegraphics` is how you end up with axis labels nobody can
read.

**Keep the appearance in one place.** The figure and the table both turn `naive`
into something a reader sees, so the label and the color are decided once, in
`code/setup.R`, which every script in the project sources:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/setup.R", from = "^method_labels",
           to = "^colors", filename = "code/setup.R (excerpt)")
```

rather than being buried in two separate `ggplot` calls that can drift apart.

::: {.callout-note}
## A sourced file has to be declared

Look at what that costs in the Snakefile: `code/setup.R` appears in the `input:`
of every rule whose script sources it. That is not decoration. Snakemake has no
way to know that `figure.R` reads another file, and an undeclared dependency is
the one failure mode it cannot warn you about --- change a color, and the figure
that uses it stays quietly stale (@sec-debugging).

The rule of thumb is exactly that condition: **declare it wherever changing it
could change an output.**

It is worth knowing what you are buying, though. `setup.R` here holds the
simulation functions as well as the plot colors, and it is therefore an input to
`rule sim` too --- so editing one hex code marks all eighteen simulations stale
and they run again. Snakemake cannot tell which half of the file you touched. On a
grid that takes seconds this is invisible; on one that takes days it is a reason to
split the file, so that each rule depends only on the part it actually uses.

That splitting has a natural end point --- shared values in one script, functions
in an `R/` directory one per file, which is the layout of an R package --- but it
is a question about project organization rather than about Snakemake, and
[IowaBiostat/project-workflow](https://github.com/IowaBiostat/project-workflow)
covers it properly: a talk plus a small example project with `R/`, `man/`,
`scripts/` and a Snakefile, worth cloning and pulling apart.
:::

## Tables

Tables are the same idea with a different output format: the script writes a LaTeX
fragment that the manuscript `\input`s.

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/code/table.R", from = "^## escape = FALSE",
           to = "^  add_header_above", filename = "code/table.R (excerpt)")
```

which produces:

```{r}
#| echo: false
#| output: asis
embed_file("transcripts/table1.tex", lang = "latex",
           filename = "out/table1.tex")
```

Points worth noting:

- The file contains a **fragment**, not a document --- no `\documentclass`, no
  `\begin{document}`. It is meant to be `\input` into something else.
- The **caption and label live with the numbers**, not in the manuscript. That way
  `\ref{tab:coverage}` resolves and the caption cannot describe a table that the
  code no longer produces.
- `escape = FALSE` lets `$n$` through as math. The cost is that everything else in
  the table must then be valid LaTeX --- a `%` or `_` in a variable name will break
  the build, which is a reasonable trade when you control the contents.

Writing `.tex` rather than pasting numbers into the manuscript by hand is the entire
argument of this book, applied to tables. A number in your paper should be
traceable to the code that computed it, and the only way to guarantee that is for
no human hand to touch it in between.

## rule all as the manifest

With figures and tables in place, `rule all` becomes the list of everything the
paper is made of:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule all:", to = '^        "build/submission.pdf"',
           filename = "Snakefile (excerpt)")
```

Here it names only the two PDFs, because everything else is reachable through them.
In a larger project --- the one in @sec-scaling has sixteen figures and tables ---
it is common to list every product explicitly instead:

```python
rule all:
    input:
        "build/paper.pdf",
        "out/figure1.pdf",
        "out/figure2.pdf",
        ...
```

This is slightly redundant, and it is what you want anyway. It gives you one place
to read off what the project produces, and it means `snakemake` still builds a
figure you have temporarily commented out of the manuscript while you rearrange
the results section.

## Try it

1. Change the color palette in `code/setup.R` and run `snakemake -n`. How many
   jobs are planned? Is that what you expected?
2. Add a `figure2` rule showing coverage against `rho` instead of `n`, and add it
   to `rule all`.
3. Delete `out/table1.tex` and run `snakemake --cores 4`. How far back does
   Snakemake go?
