# Parallelism, logs and benchmarks {#sec-parallel}

```{r}
#| include: false
source("_common.R")
```

Once your work is expressed as a DAG, you get the
parallelism for free --- Snakemake already knows which jobs do not depend on each
other.

## --cores

```bash
snakemake --cores 4
```

With eighteen independent simulations and four cores, four run, and as each
finishes another starts, until all eighteen are done and the figure can go. There is
nothing to configure --- the ordering falls out of the DAG.

`--cores all` uses every core you have. On a laptop you are also using for other
things, leave one or two free; on a shared server, be a good citizen and pick a reasonable
number.

::: {.callout-note}
## Two kinds of parallel

`--cores` controls how many *jobs* run at once, and that is usually the only kind
of parallelism you need for a simulation study. It is unrelated to parallelism
*inside* a job (`mclapply`, `future`, a multithreaded BLAS). If a rule does use
threads internally, tell Snakemake with `threads:` so its accounting stays honest:

```python
rule sim:
    threads: 4
    shell: "Rscript sim.R --cores {threads} ..."
```

Then `--cores 8` runs two such jobs at a time rather than eight.

Sometimes if each iteration is very expensive, you may want to parallelize over iterations. This sort of mechanism can be found in [University of Iowa
Biostatistics HPC tutorial](https://iowabiostat.github.io/hpc/).
:::



## Logs

Once eighteen jobs are interleaving their output, the terminal becomes unreadable,
and the run you actually want to is buried. The `log:`
directive gives each job its own file:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^rule sim:",
           to = '^        "--out \\{output\\} > \\{log\\} 2>&1"',
           filename = "Snakefile (excerpt)")
```

Two details. The log file is *declared* by `log:` but nothing writes to it
automatically --- the redirection `> {log} 2>&1` at the end of the shell command is
what does it (`2>&1` catches errors as well as normal output). And the log path
carries the same wildcards as the output, so each job gets its own file rather than
eighteen jobs overwriting one.

Anything the script prints lands there:

```{r}
#| echo: false
#| output: asis
transcript("04-log.txt")
```

Which is worth designing for. A script that prints its parameters, its sample size
after exclusions, and a summary of what it computed gives you a log directory that
answers most questions without rerunning anything. @sec-analysis leans on this
heavily.

Logs also survive failure. When a job dies, Snakemake tells you which log to read,
and the R error message is sitting in it --- as opposed to being lost in the
scrollback of a parallel run.

::: {.callout-important}
## Snakemake will not delete a log file

Outputs of a failed job are removed so that a half-written file is never mistaken
for a real one. Logs are deliberately kept, because they are the evidence. This also
means a log from a run sticks around until the job runs again.
:::

## Benchmarks

`benchmark:` records how long a job took:

```python
rule sim:
    benchmark: "benchmarks/sim/n{n}_rho{rho}_{method}.tsv"
```

```{r}
#| echo: false
#| output: asis
transcript("04-benchmark.txt")
```

The first column is seconds. This is how you find out that one corner of your
parameter grid takes forty times as long as the rest, which is worth knowing before
you scale the whole grid up by a factor of ten.

## Interrupting and resuming

Snakemake is designed to handle being stopped.

```{.default}
Terminating processes on user request, this might take some time.
```

What survives is everything that had already *finished*. Those outputs stay on
disk, and the next run does not rebuild them; it starts from wherever the
interruption left the pipeline. So the cost of stopping a twelve-hour run after
four hours is only the handful of jobs that were in flight at that moment. 

The jobs that were in flight need one more thought, because a killed job may have
got as far as writing part of its output. Snakemake tracks that, and rather than
letting a truncated file pass for a real result it stops the next run:

```{.default}
IncompleteFilesException:
The files below seem to be incomplete. …
Incomplete files:
rds/200/sim/n400_rho80_split.rds
```

The fix is to say so explicitly:

```bash
snakemake --cores 4 --rerun-incomplete
```

which discards the suspect files and rebuilds them. 

## Try it

1. In `examples/03-config`, run with `--cores 1`, then `--cores 4`, and compare.
   Why is the speedup less than four?
3. Add a `log:` directive to the `figure` rule in `examples/03-config` and check
   that `ggsave`'s message ends up in it.
