# Building the manuscript {#sec-manuscript}

```{r}
#| include: false
source("_common.R")
```

Here is the situation this chapter is about. You are using an editor that rebuilds
your PDF every time you save, so the document always matches its LaTeX source. It
does not match its *figures*: `out/figure1.pdf` was made by some script at some
point, and if you change the simulation nothing tells the manuscript about it. The
document is only as current as the last time you remembered to rerun things.

Making the manuscript a rule closes that gap.

## The rule

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^## The manuscript depends",
           filename = "Snakefile (excerpt)")
```

`latexmk` is doing the LaTeX-specific work --- running `pdflatex` and BibTeX as many
times as it takes for references and citations to settle. Snakemake's contribution
is the `input:` block: the PDF depends on the figure and the table, which depend on
the simulations, which depend on the code. One command now walks the entire chain.

The wildcard `{doc}` builds either document with the same rule, constrained to the
two names that exist. `wildcard_constraints` is attached to this rule rather than
declared at the top of the file, since it is only meaningful here.

## Separating layout from content

Both documents contain the same words. Neither contains any of them:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/paper.tex", filename = "paper.tex")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/submission.tex", filename = "submission.tex")
```

The prose is in `main.tex`, which both wrappers `\input`. This costs you nothing on
the first draft and saves you a great deal on the fourth, when the paper is going
to a second journal with a different class file, and again when you post it to
arXiv. You change the wrapper. You never touch a sentence.

Real projects extend this in the obvious way --- a wrapper per journal, one for
arXiv, one for the version with the supplement bound in --- each a rule producing
its own PDF, all sharing `main.tex`. The `man-avg-cov` project this example is
modeled on has one for arXiv and one for Biometrics, the latter of which also
splits the compiled PDF into a main document and a supplement with `qpdf`, because
that journal requires them as separate uploads.

## The cascade

This is the payoff, so it is worth watching happen. Everything is currently built
and up to date. Now change one line in the simulation code:

```{r}
#| echo: false
#| output: asis
transcript("04-cascade.txt")
```

Twenty-three jobs: eighteen simulations, the figure, the table, and both PDFs. In
the right order, with no decisions from you, and --- the part that matters --- with
no possibility of a version of the paper containing a figure from before the change.

Compare the alternative, which is remembering to run four things in order and being
right about it every time, for the entire life of the project.

## Making LaTeX quiet

Two flags in that rule are there for the sake of your terminal:

`-interaction=nonstopmode` stops LaTeX from halting at an error to ask you what to
do. In a pipeline, an interactive prompt is a hang, not a question.

`> {log} 2>&1` sends the several hundred lines LaTeX produces to `logs/paper-latexmk.log`.
When the build fails, the error is in there, and `latexmk` also leaves the more
readable `build/paper.log`.

::: {.callout-warning}
## LaTeX's own idea of what is up to date

`latexmk` does its own dependency tracking and will sometimes decide there is
nothing to do even when Snakemake has just asked it to rebuild. If a rerun leaves
the PDF unchanged when you expected otherwise, delete the `build/` directory and
try again --- and note that this is exactly why the compiled PDF should not be the
only place a result appears.
:::

## What goes in git

The `.gitignore` for a project of this shape:

```{.default filename=".gitignore"}
.snakemake/
rds/
out/
logs/
benchmarks/
build/
```

Everything ignored is regenerable, and the claim that it is regenerable is now
enforced by a program rather than by your recollection. Commit the code, the
`.tex` files, and `config.yaml` --- and note that the compiled PDF is on the
ignored side of that line too. It is an output like any other.

The awkward case is a simulation that takes three days, which colleagues who clone
the repository cannot rebuild over coffee. The answer is still not to commit it.
Send them `rds/`, or keep it somewhere outside the repository that the Snakefile
knows how to find; either way it travels beside the repository rather than inside
it. This is the payoff for keeping `rds/` separate from `out/` --- you hand over
the expensive half on its own, and every figure, table and PDF rebuilds from it in
seconds. @sec-scaling covers the mechanics.

## Try it

1. Change the figure's y-axis label and run `snakemake --cores 4`. Which jobs run?
   Which do not, and why not?
2. Add a `supplement.tex` wrapper that includes only the table, plus a rule to
   build it, and add it to `rule all`.
3. Introduce a deliberate LaTeX error in `main.tex` and run the build. Where does
   Snakemake tell you to look, and what is in that file?
