# Building a manuscript {#sec-manuscript}

```{r}
#| include: false
source("_common.R")
```

Move to `examples/04-manuscript`. It is the simulation study of the last three
chapters with two things added: rules that turn the eighteen result files into a
figure and a table, and a rule that compiles a paper out of them.

Here is the situation this chapter is about. You are using an editor that rebuilds
your PDF every time you save, so the document always matches its LaTeX source. It
does not match its *figures*: `out/figure1.pdf` was made by some script at some
point, and if you change the simulation nothing tells the manuscript about it. The
document is only as current as the last time you remembered to rerun things.

Making the manuscript a rule closes that gap.

## The rule

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/Snakefile", from = "^## The manuscript depends",
           filename = "Snakefile (excerpt)")
```

`latexmk` is doing the LaTeX-specific work --- running `pdflatex` and BibTeX as many
times as it takes for references and citations to settle. Snakemake's contribution
is the `input:` block: the PDF depends on the figure and the table, which depend on
the simulations, which depend on the code. One command now walks the entire chain.

Written out, the rules make this shape:

```{mermaid}
%%| fig-align: center
flowchart LR
  sim(sim) --> figure1(figure1)
  sim --> table1(table1)
  figure1 --> manuscript(manuscript)
  table1 --> manuscript
  manuscript --> allrule(all)
```

Every arrow is a line in the Snakefile, and every one of them is a claim that
Snakemake will enforce. Nothing reaches `build/paper.pdf` except through a rule.

That diagram is not hand-drawn: it is the graph Snakemake reports for this
workflow, and you can get one for your own with

```bash
snakemake --rulegraph | dot -Tpdf > rules.pdf
```

One node per *rule*, which is the readable view. `snakemake --dag` draws one node
per *job* instead, so `sim` becomes the eighteen separate boxes it really is
(@sec-debugging).

`paper.tex` itself is only a wrapper --- a document class, `\input{preamble.tex}`, and
`\input{main.tex}` --- so that the prose lives in one file and the formatting in
another:

```{r}
#| echo: false
#| output: asis
embed_file("examples/04-manuscript/paper.tex", filename = "paper.tex")
```

::: {.callout-note}
## Journal versions, arXiv, supplements

A real paper is eventually built more than one way. The
separation above is what makes multiple versions with different formats lightweight --- each variant is another wrapper
around the same `main.tex`, and another rule producing its own PDF.

Setting that up well is a LaTeX question rather than a Snakemake one, and
[pbreheny/manuscript-template](https://github.com/pbreheny/manuscript-template)
already covers it.
:::

## The cascade

This is the payoff, so it is worth watching happen. Everything is currently built
and up to date. Now change one line in the simulation code:

```{r}
#| echo: false
#| output: asis
transcript("04-cascade.txt")
```

Twenty-two jobs: eighteen simulations, the figure, the table and the PDF. In the
right order, with no decisions from you, and --- the part that matters --- with no
possibility of a version of the paper containing a figure from before the change (provided you've specified the Snakefile completely).

Compare the alternative, which is remembering to run four things in order and being
right about it every time, for the entire life of the project (and with the number being much greater than four as the manuscript grows).

::: {.callout-warning}
## LaTeX's own idea of what is up to date

`latexmk` does its own dependency tracking and will sometimes decide there is
nothing to do even when Snakemake has just asked it to rebuild. If a rerun leaves
the PDF unchanged when you expected otherwise, delete the `build/` directory and
try again.
:::

## Try it

1. Change the figure's y-axis label and run `snakemake --cores 4`. Which jobs run?
   Which do not, and why not?
