# A data analysis pipeline {#sec-analysis}

```{r}
#| include: false
source("_common.R")
```

Many biostatisticians spend more time on analyses than on simulation studies, and
the shape of the work is different: instead of a fan-out over a parameter grid, a
mostly linear chain from a file someone emailed you to a report you send back. The
Snakemake ideas transfer directly..

Move to `examples/05-analysis`. It is a small randomized trial: 200 participants,
two arms, FEV1 measured at baseline and week 12.

## The pipeline

```{r}
#| echo: false
#| output: asis
transcript("05-job-stats.txt")
```

```{mermaid}
%%| fig-align: center
flowchart LR
  clean(clean) --> table1(table1)
  clean --> analysis(analysis)
  clean --> figure1(figure1)
  clean --> report(report)
  analysis --> table2(table2)
  table1 --> report
  table2 --> report
  figure1 --> report
  report --> allrule(all)
```

`clean` feeds nearly everything, because every product describes the same analysis
set. `table1` says who was studied and `table2` what happened to them, which is why
one hangs off the cleaned data and the other off the models.

## Cleaning once

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/scripts/clean.R", from = "^raw <- read.csv",
           to = "^log_step\\(d, \"with a week 12", filename = "scripts/clean.R (excerpt)")
```

Every decision about the data is in this one script, and it announces each one:

```{r}
#| echo: false
#| output: asis
transcript("05-clean-log.txt")
```

That is the participant flow for your CONSORT diagram, produced as a side effect of
the pipeline rather than counted by hand, and regenerated every time the data
changes. When the coordinator sends an updated export in June, `snakemake` reruns
everything downstream and this log tells you immediately whether the exclusions
still look the way they did in March.

The alternative --- each analysis script starting with its own `read.csv()` and its
own quiet decisions about who to drop --- is how two tables in the same report end
up with different denominators.

## The report is a view, not a computation

Here is the part specific to analysis work. It is normal to put everything in one
`.qmd` file: read the data, fit the models, print the tables, write the text. It
works, and it has two failure modes that get worse as the project grows.

The first is time. A report that refits everything takes as long as the slowest
model *every time you fix a typo*, and the temptation is to start caching chunks,
which I've had very little luck with.

The second is scope. Rendering becomes all-or-nothing --- you cannot ask for just
the models --- and the results exist only inside the rendered HTML, so the next
project that needs them starts by copying and pasting.

Splitting it fixes both. The scripts compute and save; the report reads and
displays:

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/report.qmd", from = "^```\\{r\\}$",
           to = "^result\\.$", filename = "report.qmd (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/Snakefile", from = "^## Listing every object",
           filename = "Snakefile (excerpt)")
```

Listing each object the report reads is what makes the report trustworthy. Rendering
it cannot use a stale table, because Snakemake will rebuild the table first. And
the fitted models are in `rds/analysis.rds`, so the manuscript you write next year
can load them without rerunning anything.

Everything the report needs to say in prose comes through inline R:

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/report.qmd",
           from = "^The estimated treatment effect is",
           to = "^model, and is insensitive", filename = "report.qmd (excerpt)")
```

so the sentence and the table cannot disagree.

## Models as data

This of the models as a step in the pipeline where the input is data and the output is the model results (a different type of data) that can serve as input into the next step. Each model should get its own file.

[To be continued...]

## When the data changes

Which it will, usually the week after you send the report. Drop in the new data
and:

```bash
snakemake --cores 4
```

Cleaning reruns, the models refit, Table 1 and the figure regenerate, the report
re-renders, and the participant-flow log tells you what changed about the analysis
set. The work of a data update goes from an afternoon of retracing your steps to one
command and a careful look at one log file.

## Try it

1. Edit a value in `data/trial.csv` and run `snakemake -n`. What reruns?
3. Delete `rds/clean.rds` but not the things built from it. What does Snakemake
   do, and is that what you expected?
