# A data analysis pipeline {#sec-analysis}

```{r}
#| include: false
source("_common.R")
```

Most biostatisticians spend more time on analyses than on simulation studies, and
the shape of the work is different: instead of a fan-out over a parameter grid, a
mostly linear chain from a file someone emailed you to a report you send back. The
Snakemake ideas transfer directly, and one of them --- the report as a rule ---
addresses a problem specific to this kind of work.

Move to `examples/05-analysis`. It is a small randomized trial: 200 participants,
two arms, FEV1 measured at baseline and week 12.

## The pipeline

```{r}
#| echo: false
#| output: asis
transcript("05-job-stats.txt")
```

```{mermaid}
%%| fig-align: center
flowchart LR
  R[data/trial.csv] --> C[rds/analysis.rds]
  C --> M[rds/models.rds]
  C --> T[rds/table1.rds]
  C --> F[out/figure1.png]
  M --> P[report.html]
  T --> P
  F --> P
  C --> P
```

## Raw data is read-only

The first rule reads `data/trial.csv` and never writes to it. That is not a
Snakemake requirement; it is the convention that makes everything else work, and it
deserves stating plainly: **the raw file stays exactly as it arrived.** Every
correction, exclusion and recoding happens in code, downstream, where it can be read
and argued with.

Note also what is *not* in the Snakefile. `data/make-raw-data.R`, which generated
the CSV for this example, is not a rule. If it were, Snakemake would treat the raw
data as something it could regenerate, and would do so --- silently discarding the
real file the moment it decided it was out of date. Anything you cannot afford to
lose should be an input to your pipeline and never an output of it.

## Cleaning once

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/scripts/clean.R", from = "^raw <- read.csv",
           to = "^log_step\\(d, \"with a week 12", filename = "scripts/clean.R (excerpt)")
```

Every decision about the data is in this one script, and it announces each one:

```{r}
#| echo: false
#| output: asis
transcript("05-clean-log.txt")
```

That is the participant flow for your CONSORT diagram, produced as a side effect of
the pipeline rather than counted by hand, and regenerated every time the data
changes. When the coordinator sends an updated export in June, `snakemake` reruns
everything downstream and this log tells you immediately whether the exclusions
still look the way they did in March.

The alternative --- each analysis script starting with its own `read.csv()` and its
own quiet decisions about who to drop --- is how two tables in the same report end
up with different denominators.

## The report is a view, not a computation

Here is the part specific to analysis work. It is normal to put everything in one
`.qmd` file: read the data, fit the models, print the tables, write the text. It
works, and it has two failure modes that get worse as the project grows.

The first is time. A report that refits everything takes as long as the slowest
model *every time you fix a typo*, and the temptation is to start caching chunks,
at which point you no longer know whether the numbers in the output correspond to
the code in the file.

The second is scope. Rendering becomes all-or-nothing --- you cannot ask for just
the models --- and the results exist only inside the rendered HTML, so the next
project that needs them starts by copying and pasting.

Splitting it fixes both. The scripts compute and save; the report reads and
displays:

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/report.qmd", from = "^```\\{r\\}$",
           to = "^result\\.$", filename = "report.qmd (excerpt)")
```

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/Snakefile", from = "^## Listing every object",
           filename = "Snakefile (excerpt)")
```

Listing each object the report reads is what makes the report trustworthy. Rendering
it cannot use a stale table, because Snakemake will rebuild the table first. And
the models are in `rds/models.rds`, so the manuscript you write next year can
load them without rerunning anything.

Everything the report needs to say in prose comes through inline R:

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/report.qmd",
           from = "^The estimated treatment effect is$",
           to = "^in the adjusted model", filename = "report.qmd (excerpt)")
```

so the sentence and the table cannot disagree.

## Models as data

```{r}
#| echo: false
#| output: asis
embed_file("examples/05-analysis/scripts/model.R", from = "^specs <- list",
           to = "^\\}\\)\\)", filename = "scripts/model.R (excerpt)")
```

Two things worth copying here. The model specifications are a named list, so adding
a sensitivity analysis means adding one line and touching nothing else. And the
script saves both the fitted objects and a tidy data frame of the estimand --- the
first for anything you did not anticipate, the second so the report does not need to
know how to dig a confidence interval out of an `lm`.

```{r}
#| echo: false
#| output: asis
transcript("05-model-log.txt")
```

## When the data changes

Which it will, usually the week after you send the report. Drop in the new export
and:

```bash
snakemake --cores 4
```

Cleaning reruns, the models refit, Table 1 and the figure regenerate, the report
re-renders, and the participant-flow log tells you what changed about the analysis
set. The work of a data update goes from an afternoon of retracing your steps to one
command and a careful look at one log file.

## Try it

1. Edit a value in `data/trial.csv` and run `snakemake -n`. What reruns?
2. Add a rule producing a subgroup analysis by smoking status, and show it in the
   report.
3. Delete `rds/analysis.rds` but not the models. What does Snakemake do, and is
   that what you expected?
