# When things go wrong {#sec-debugging}

```{r}
#| include: false
source("_common.R")
```

Most of the trouble people have with Snakemake comes from one of two questions:
*why is it doing that?* and *why is it not doing that?* This chapter provides some useful tools for debugging.

## Look before you run

`-n` is the answer to a surprising fraction of problems. It builds the DAG, decides
what it would do, and does none of it:

```bash
snakemake -n              # what would you do?
```

Each planned job comes with a `reason:` line:

```{r}
#| echo: false
#| output: asis
transcript("04-reason.txt")
```

## What is the status of each file?

`--summary` lets you inspect the files statuses, one row per file Snakemake knows about:

```{r}
#| echo: false
#| output: asis
transcript("09-summary.txt")
```

The `plan` column tells you what would happen and the `status` column whether the
recorded provenance still matches.

## Making it rerun anyway

| Command | What it does |
|---|---|
| `snakemake -R sim` | rerun rule `sim` and everything downstream |
| `snakemake --forcerun sim` | same thing, spelled out |
| `snakemake -f out/figure1.pdf` | rebuild just that file |
| `snakemake -F` | force everything, from scratch |
| `snakemake --delete-all-output` | delete everything the workflow produced |

`-R` is what you can use when you say "I changed something
Snakemake cannot see". E.g. a package upgrade (more on this in the future).

## Making it not rerun

The mirror image: files are current but Snakemake wants to rebuild them, usually
because they were copied from somewhere or restored from a backup.

```bash
snakemake --touch
```

`--touch` marks every output as up to date without running anything. Use it when you
know the files are fine, and note that you are making an assertion Snakemake cannot
check.

The case where you will actually want it is results that arrived from a
collaborator without the metadata that explains them (`.snakefile`), and there `--touch` on its
own is not enough --- @sec-sharing has the recipe.

`--rerun-triggers` narrows what Snakemake is allowed to notice. All five signals
are on by default:

```bash
snakemake --rerun-triggers code input mtime params software-env   # the default
snakemake --rerun-triggers input                                  # only file contents
```

which is occasionally useful.

[to be continued...]

<!-- ::: {.callout-note} -->
<!-- ## `--rerun-triggers mtime` is not `make` -->

<!-- Snakemake's own help text offers `--rerun-triggers mtime` as "the traditional way -->
<!-- of just considering file modification dates". It does not behave that way: on -->
<!-- 9.13, `touch`ing an input under `--rerun-triggers mtime` still does not cause a -->
<!-- rerun, because the checksum comparison described in @sec-rerun applies -->
<!-- regardless. If you want a rebuild, ask for one with `-R` rather than trying to -->
<!-- manufacture one through timestamps. -->
<!-- ::: -->

## The errors you will see

### No rule to produce

```{r}
#| echo: false
#| output: asis
transcript("09-missing-rule.txt")
```

You asked for a file no rule can make. Nine times in ten it is a typo, in the target
or in a rule's `input:`. Check spelling, check the directory, and check whether wildcards are being read correctly.

[Details on debugging wildcards.]

### A job failed

```{r}
#| echo: false
#| output: asis
transcript("09-job-failed.txt")
```

Snakemake reports the rule, the wildcards, and the exact shell command. **Copy that
command and run it yourself** --- you will get the R error directly, and you can
step through the script interactively with the same arguments.

Two things happened that you did not see. Any output the job had already written
was deleted --- you get `Removing output files of failed job … since they might be
corrupted`, so a half-written `.rds` can never be mistaken for a real result. (Here
the script died before writing anything, so there was nothing to remove.) The log
file, by contrast, was kept: if the rule declared one, the R error message is in it so it is often a good idea to start there.

The example above is the wildcard problem from @sec-wildcards: `bayes` matched
`{method}`, so Snakemake built a perfectly good job for a method that does not
exist. `wildcard_constraints` can turn this into a "no rule to produce" error before
anything runs.

### Incomplete files

> The files below seem to be incomplete.

A previous run was stopped while a job was part-way through writing its output.

`snakemake --rerun-incomplete` discards the suspect files and rebuilds them.

## Habits that prevent most of this

1. **List every script as an input.** The one failure mode Snakemake cannot warn
   you about is the dependency you did not declare.
2. **Dry-run before long runs.** A minute may save you hours.
3. **Then smoke-test with a small config.** `--config iterations=20` finds broken
   plumbing without simulating anything.
4. **Give every rule a log.** Debugging a parallel run without logs can be a headache.
5. **Delete `rds/` and `out/` occasionally and rebuild from scratch.** It is the
   only way
   to know the pipeline still works end to end --- and the thing you are implicitly
   promising anyone who reads your paper.
