# When things go wrong {#sec-debugging}

```{r}
#| include: false
source("_common.R")
```

Most of the trouble people have with Snakemake comes from one of two questions:
*why is it doing that?* and *why is it not doing that?* This chapter is a reference
for both, plus the errors you will actually see.

## Look before you run

`-n` is the answer to a surprising fraction of problems. It builds the DAG, decides
what it would do, and does none of it:

```bash
snakemake -n              # what would you do?
snakemake -n --quiet      # just the summary table
snakemake -p              # print the shell commands as they run
```

Each planned job comes with a `reason:` line:

```{r}
#| echo: false
#| output: asis
transcript("04-reason.txt")
```

::: {.callout-note}
## `-r`/`--reason` no longer exists

Reasons are printed by default now, and the flag was removed --- on 9.13 it is a
hard error, not a deprecation warning. If you find a tutorial recommending
`snakemake -n -r`, it predates version 8, and whatever else it tells you about
cluster execution is likely out of date too.
:::

## Why is it rebuilding?

@sec-rerun covers what makes an output stale; `--summary` is how you inspect it
after the fact, one row per file Snakemake knows about:

```{r}
#| echo: false
#| output: asis
transcript("09-summary.txt")
```

The `plan` column tells you what would happen and the `status` column whether the
recorded provenance still matches.

The usual surprises:

**A whole grid reruns after a cosmetic edit.** You changed a comment in `sim.R`, and
`sim.R` is an input to eighteen jobs. Snakemake compares content, so *any* change
counts, including whitespace. There is no way to say "this edit does not matter";
if you are sure, `--touch` (below) is the escape hatch.

**A rule reruns after you edited its shell command.** Expected, and desirable ---
the command is part of the provenance.

**Nothing reruns after you edit a script that is not listed as an input.** This one
is silent and it is the dangerous one. If a rule calls `Rscript scripts/figure.R`
but only lists `{input.rds}`, then editing `figure.R` changes nothing as far as
Snakemake is concerned, and your figure quietly stays stale. **List your scripts as
inputs**, and whatever they source (@sec-rerun).

## Making it rerun anyway

| Command | What it does |
|---|---|
| `snakemake -R sim` | rerun rule `sim` and everything downstream |
| `snakemake --forcerun sim` | same thing, spelled out |
| `snakemake -f out/figure1.pdf` | rebuild just that file |
| `snakemake -F` | force everything, from scratch |
| `snakemake --delete-all-output` | delete everything the workflow produced |

`-R` is the one you will use. It is the right response to "I changed something
Snakemake cannot see" --- a package upgrade, an edit to a script you forgot to list,
a change to a data file that lives outside the project.

## Making it stop rerunning

The mirror image: files are current but Snakemake wants to rebuild them, usually
because they were copied from somewhere or restored from a backup.

```bash
snakemake --touch
```

`--touch` marks every output as up to date without running anything. Use it when you
know the files are fine, and note that you are making an assertion Snakemake cannot
check.

`--rerun-triggers` narrows what Snakemake is allowed to notice. All five signals
are on by default:

```bash
snakemake --rerun-triggers code input mtime params software-env   # the default
snakemake --rerun-triggers input                                  # only file contents
```

which is occasionally useful --- dropping `params`, say, when you are sweeping a
setting that does not really affect the output and do not want the whole grid
rebuilding each time.

::: {.callout-note}
## `--rerun-triggers mtime` is not `make`

Snakemake's own help text offers `--rerun-triggers mtime` as "the traditional way
of just considering file modification dates". It does not behave that way: on
9.13, `touch`ing an input under `--rerun-triggers mtime` still does not cause a
rerun, because the checksum comparison described in @sec-rerun applies
regardless. If you want a rebuild, ask for one with `-R` rather than trying to
manufacture one through timestamps.
:::

## The errors you will see

### No rule to produce

```{r}
#| echo: false
#| output: asis
transcript("09-missing-rule.txt")
```

You asked for a file no rule can make. Nine times in ten it is a typo, in the target
or in a rule's `input:`. Check spelling, check the directory, and check whether a
`wildcard_constraints` is stopping the match --- a constraint of `\d+` on `{n}` will
make `nfifty_rho0_naive.rds` unmatchable, which is the point, but the message does
not say so.

### A job failed

```{r}
#| echo: false
#| output: asis
transcript("09-job-failed.txt")
```

Snakemake reports the rule, the wildcards, and the exact shell command. **Copy that
command and run it yourself** --- you will get the R error directly, and you can
step through the script interactively with the same arguments.

Two things happened that you did not see. Any output the job had already written
was deleted --- you get `Removing output files of failed job … since they might be
corrupted`, so a half-written `.rds` can never be mistaken for a real result. (Here
the script died before writing anything, so there was nothing to remove.) The log
file, by contrast, was kept: if the rule declared one, the R error message is in it.

The example above is the wildcard problem from @sec-wildcards: `bayes` matched
`{method}`, so Snakemake built a perfectly good job for a method that does not
exist. `wildcard_constraints` turns this into a "no rule to produce" error before
anything runs.

### Incomplete files

> The files below seem to be incomplete.

A previous run was stopped while a job was part-way through writing its output ---
by `Ctrl-C` as readily as by a hard kill, since `Ctrl-C` terminates running jobs
rather than waiting for them (@sec-parallel). `snakemake --rerun-incomplete`
discards the suspect files and rebuilds them.

### Directory cannot be locked

> Error: Directory cannot be locked.

Either another Snakemake is running in this directory --- check before you assume it
is not --- or one was killed without cleaning up. If you are sure nothing is
running, `snakemake --unlock`.

### The ILP solver warning

> Failed to solve scheduling problem with ILP solver, falling back to greedy
> scheduler.

Harmless. Snakemake schedules jobs optimally when it can find a working solver, and
falls back to a greedy scheduler when it cannot; the workflow runs correctly either
way. It shows up on some Homebrew and conda installs. Ignore it, or install
`pulp` properly if the noise bothers you.

## Reading a workflow you did not write

```{r}
#| echo: false
#| output: asis
transcript("09-list.txt")
```

`--list` shows the rule names. Then:

```bash
snakemake --dag | dot -Tpdf > dag.pdf          # the job graph, one node per job
snakemake --rulegraph | dot -Tpdf > rules.pdf  # the rule graph, one node per rule
```

Both need [Graphviz](https://graphviz.org/) for `dot`. For a grid of any size, use
`--rulegraph`: `--dag` draws a node per *job*, which for the example in
@sec-manuscript means twenty-two of them.

Failing that, read `rule all` first. It tells you what the project is for, and
everything else follows from it.

## Habits that prevent most of this

1. **List every script as an input.** The one failure mode Snakemake cannot warn
   you about is the dependency you did not declare.
2. **Dry-run before long runs.** Twenty seconds against three hours.
3. **Smoke-test with a small config.** `--config iterations=20` finds broken
   plumbing without simulating anything.
4. **Give every rule a log.** Debugging a parallel run without logs is guesswork.
5. **Delete `rds/` and `out/` occasionally and rebuild from scratch.** It is the
   only way
   to know the pipeline still works end to end --- and the thing you are implicitly
   promising anyone who reads your paper.
