# Sharing results {#sec-sharing}

```{r}
#| include: false
source("_common.R")
```

Everything so far has assumed one machine. A coauthor who clones your repository
gets the code, which is the cheap half of the project; what they do not get is the
two days of compute that produced the numbers. This chapter is about handing over
the expensive half, and about the one thing that goes wrong when you do.

## The expensive half and the cheap half

This is where the discipline of @sec-shape --- computed results in `rds/`, figures
and tables in `out/` --- stops being tidiness and starts paying rent. What you
share is `rds/`, on its own: it is the expensive half, and a collaborator who has
it can regenerate every figure and table in seconds.

A simulation that takes two days to run is not a build artifact in the way a figure
is. 

## Why not git

Git is the tool you would
reach for otherwise, except that git is bad at exactly this kind of content. A few
hundred megabytes of `.rds` files bloats the repository permanently --- history
keeps every version you ever committed, so it is not fixed by deleting them later
--- diff and merge do nothing useful on serialized objects, and results derived
from protected health information have no business on GitHub at all.

## dat

[`dat`](https://github.com/pbreheny/dat) is built for the gap the previous section
described. It gives a directory a git-style push/pull workflow backed by AWS S3,
without the version history:

```bash
cd rds
dat init
dat push
```

and, on a collaborator's machine:

```bash
dat clone jsmith7.rds.bucket rds
dat pull
```

What makes it more than a wrapper around `rsync` is that it detects conflicts: if
someone has modified a file you are about to overwrite, it says so rather than
silently clobbering their work. It hashes file contents, so pushing a directory in
which one of four hundred simulations changed transfers one file, and `dat status`
will tell you whether you are in sync.

The division of labor is natural. Git holds the code: small, textual, and wanting
history. `dat` holds the results: large, binary, and only ever wanting to be
*current*. Each is one command to push, and a collaborator with both has your
entire project.

## What does not arrive with the files

Sharing tool aside, there are complexities with snakemake when sharing result files.

Your coauthor clones the repository and pulls the results folder. Every `.rds` file
is sitting there. But Snakemake's record of how those files were made ---
`.snakemake/`, which is local and was never shared (@sec-rerun):

```{r}
#| echo: false
#| output: asis
transcript("10-no-metadata.txt")
```

Twenty-two jobs, eighteen of them the simulations that just arrived. On a real
project that is the two days of compute the sharing was supposed to avoid.

### Why bare `--touch` is not the fix

`--touch` marks outputs as up to date without running anything, which sounds like
exactly what is needed. Run it as it comes and it does too much:

```{r}
#| echo: false
#| output: asis
transcript("10-touch-everything.txt")
```

Twenty-two of twenty-two. It asserted the figure, the table and the paper along
with the simulations are up to date, and there is now nothing left for Snakemake to do. Those are
precisely the things you wanted rebuilt: they cost seconds, and rebuilding them is
what makes the paper on your disk yours rather than a copy of your coauthor's.

### The recipe: touch the results, build the rest

Be specific. Name the expensive outputs, and leave everything downstream alone:

```{r}
#| echo: false
#| output: asis
transcript("10-touch-results.txt")
```

Eighteen touched, and the four jobs that matter are still to do. Then run
`snakemake` normally and the figure, the table and the PDF are built on your
machine, in seconds, from the results you were sent.

Read it inside out: `--summary all` prints one row per output
file the workflow knows about, `awk '{print $1}'` takes the filename column, and
`grep '\.rds$'` keeps the simulation results while dropping the figures, tables,
PDFs and the header line. `--touch` acts at the jobs to be run, so `--force` acts on the jobs you name and nothing else. 


::: {.callout-warning}
## `--touch` is an assertion, not a check

You are telling Snakemake that files it has never seen are the correct output of
the current code. It has no way to verify that, and it will believe you. If your
coauthor generated those results with a different version of `sim.R` than the one
you just cloned, the pipeline is now quietly wrong.
:::

## Try it

In `examples/04-manuscript`, with everything already built:

1. Delete `.snakemake/` and run `snakemake -n`. Which of the planned jobs are the
   expensive ones, and which would have cost you seconds anyway?
2. Run bare `snakemake --touch`, then `snakemake -n`. Confirm that nothing is left
   to do.
3. Delete `.snakemake/` again, run the `--touch --force` recipe, and then
   `snakemake`. Four jobs run. Then edit `scripts/sim.R` and confirm the full
   cascade comes back.
