# Scaling up {#sec-scaling}

```{r}
#| include: false
source("_common.R")
```

The example in @sec-manuscript is a real pipeline in miniature: eighteen simulations,
a figure, a table, two documents. A real methods paper is the same shape with the
numbers changed --- a few hundred simulation jobs, seventeen figures and tables,
results measured in days of compute rather than seconds. This chapter is about what
changes and what does not.

Almost nothing about the Snakefile changes. What changes is where things live.

## Two repositories

The natural instinct is to put everything in the manuscript's repository. That works
until the second paper, at which point you find yourself copying the estimator into
a new directory and maintaining two versions of it.

The structure that holds up splits the work in two:

```{mermaid}
%%| fig-align: center
flowchart TB
  subgraph M["manuscript repo"]
    S[Snakefile]
    T[main.tex]
    C[code/figure1.R ...]
  end
  subgraph R["methods repo (an R package)"]
    F[R/ -- the estimators]
    Q[scripts/ -- one per simulation]
  end
  S -->|runs| Q
  Q -->|writes| D[(rds/ -- cached results)]
  D --> C
  C --> T
```

**The methods repository is an R package.** `R/` holds the estimators, with the
usual `DESCRIPTION` and `NAMESPACE`, so it can be installed, tested, versioned and
cited on its own. `scripts/` holds one script per kind of simulation, each taking
`optparse` flags and writing one `.rds` named for its parameters --- exactly the
scripts of @sec-rscript, just more of them.

**The manuscript repository holds the paper**: the `.tex` files, the scripts that
draw each figure, and the Snakefile that ties it together. Its rules call the *other*
repository's scripts.

The payoff is that the estimator has one home. The next paper gets a new manuscript
repository and reuses the package; a reader who wants the method installs the
package and ignores the paper entirely.

## Pointing at the other repository

The Snakefile needs to know where the methods repository is, and coauthors keep it
in different places. The straightforward approach is a config value:

```{.yaml filename="config.yaml"}
res-loc: "~/repos/lasso-confint/"
```

```{.python filename="Snakefile"}
import os
LOC = os.path.expanduser(config["res-loc"])

rule distribution_results:
    input:
        script = f"{LOC}scripts/distribution_results.R"
    output:
        f"{LOC}rds/{ITER}/original/{{distribution}}_{{n}}_{{method}}.rds"
    shell:
        "Rscript {input.script} --n {wildcards.n} --method {wildcards.method} "
        "--loc {LOC}"
```

Two things to notice. The `f` prefix makes a Python f-string, where `{LOC}` is
substituted immediately; wildcards, which must survive to be matched later, are
written `{{n}}` to escape them. And `--loc {LOC}` passes the location to the R
script too, so that it can find the package and write its results in the right
place.

If your collaborators will not maintain a config file, you can look the path up by
username instead:

```{.python filename="Snakefile"}
import getpass
LOC = {
    "pbreheny": "~/res/lasso-confint/",
    "hlogan":   "~/repos/lasso-confint/",
}.get(getpass.getuser(), "~/res/lasso-confint/")
```

This is what `man-avg-cov` does, and it is genuinely convenient for a small group.
It also stops working the moment someone new joins, so treat it as a default that
the config can override rather than the only mechanism.

## Results outside the repository

Notice that the `.rds` files in that layout live in the *methods* repository, not
the manuscript's. This is deliberate, and it is the main structural difference from
the toy example.

A simulation that takes two days to run is not a build artifact in the way a figure
is. You do not want it deleted when you clean the manuscript directory, you want it
shared between the paper and the talk and the grant renewal, and you do not want it
in git. Keeping the cache in one place outside the manuscript --- keyed by
parameters, so a run for `n = 400` is never confused with a run for `n = 100` ---
gets you all three.

The iteration count is part of the path for the same reason:

```
rds/1000/original/laplace_autoregressive_0_100_101_gaussian_100_pipep.rds
    ^^^^
```

so the 20-iteration smoke test and the 1000-iteration final run coexist, and
switching between them is a config edit rather than an act of destruction. This is
the rule from @sec-params-or-filename at the scale where it earns its keep: were
the iteration count merely passed to the script rather than naming the directory,
a smoke test would cost you two days of compute to undo.

### Sharing the folder

This is where the discipline of @sec-shape --- computed results in `rds/`, figures
and tables in `out/` --- stops being tidiness and starts paying rent. What you
share is `rds/`, on its own: it is the expensive half, and a collaborator who has
it can regenerate every figure and table in seconds. The question is how.

Email and Dropbox both work, and both fail the same way: somebody overwrites a file
you were about to change, and the loser never finds out. Git is the tool you would
reach for otherwise, except that git is bad at exactly this kind of content. A few
hundred megabytes of `.rds` files bloats the repository permanently, diff and merge
do nothing useful on serialized objects, and results derived from protected health
information have no business on GitHub at all.

[`dat`](https://github.com/pbreheny/dat) is built for that gap. It gives a
directory a git-style push/pull workflow backed by S3, without the version history:

```bash
cd rds
dat init
dat push
```

and, on a collaborator's machine:

```bash
dat clone jsmith7.res.lasso-confint rds
dat pull
```

What makes it more than a wrapper around `rsync` is that it detects conflicts: if
someone has modified a file you are about to overwrite, it says so rather than
silently clobbering their work. It hashes file contents, so pushing a directory in
which one of four hundred simulations changed transfers one file, and `dat status`
will tell you whether you are in sync without touching the network.

The division of labor is natural. Git holds the code: small, textual, and wanting
history. `dat` holds the results: large, binary, and only ever wanting to be
*current*. Each is one command to push, and a collaborator with both has your
entire project.

### Telling Snakemake the results are fine

There is a catch, and it bites the first time you use any of this.

Your coauthor clones the repository and pulls the results folder. Every `.rds`
file is sitting there. But Snakemake's record of how those files were made ---
`.snakemake/`, which is local and was never shared --- does not exist on their
machine, so it has nothing to compare against and assumes the worst:

```{r}
#| echo: false
#| output: asis
transcript("10-no-metadata.txt")
```

Twenty-three jobs, eighteen of them the simulations that just arrived. On a real
project that is the two days of compute the sharing was supposed to avoid.

`--touch` is the escape hatch: it marks outputs as up to date without running
anything. Reaching for it on its own, though, is a coin flip. Of the two conditions
in @sec-first-snakefile only the first survives here --- there are no recorded
checksums to compare against, so the timestamp gate is all that is left --- and
`--touch` acts only on jobs it would otherwise have run. So if the incoming `.rds`
files happen to be
newer than the scripts, nothing looks out of date, nothing gets touched, and you
are left with no provenance at all --- which Snakemake mentions, quietly, at the
end of the next dry run:

```{.default filename="the branch you do not want"}
Nothing to be done (all requested files are present and up to date).
22 jobs have missing provenance/metadata so that it in part cannot be used to
trigger re-runs.
Rules with missing metadata: figure1 manuscript sim table1
```

That is a pipeline that will no longer notice when you edit `sim.R`. Which way it
goes depends on the order files happened to land on disk when you cloned and
pulled, and you have no control over that.

The fix is to stop leaving it to chance: force the expensive outputs to be
considered out of date, so the touch always has something to do.

```bash
snakemake --touch --forcerun \
  $(snakemake --summary all | awk '{print $1}' | grep '\.rds$')
```

Read the inside out: `--summary all` prints one row per output file the workflow
knows about, `awk '{print $1}'` takes the filename column, and `grep '\.rds$'`
keeps the simulation results and drops the figures, tables, PDFs and the header
line. Those filenames become the `--forcerun` list. Snakemake then walks the whole
graph, touching rather than running --- and, crucially, writing the provenance
records it was missing:

```{r}
#| echo: false
#| output: asis
transcript("10-touch-forcerun.txt")
```

From that point the project behaves normally: nothing to be done, and editing
`code/sim.R` correctly plans all twenty-three jobs again. The cheap products ---
the figure, the table, the PDF --- get touched too, so if you would rather rebuild
those from the results you were given, delete them afterwards and run `snakemake`
as usual. It takes seconds and it means the paper on your disk was compiled by
your machine from the numbers you were sent.

::: {.callout-warning}
## `--touch` is an assertion, not a check

You are telling Snakemake that files it has never seen are the correct output of
the current code. It has no way to verify that, and it will believe you. If your
coauthor generated those results with a different version of `sim.R` than the one
you just cloned, the pipeline is now quietly wrong, and nothing downstream will
ever notice.

Use it when you know the results and the code came from the same commit --- which
is an argument for tagging the code when you push results, and for keeping
`iterations` and the seed in the config where a human can read them off.
:::

## Chained rules

Expensive intermediate steps become their own rules, rather than being folded into
the script that needs them:

```{.python filename="Snakefile"}
rule fit_gam:
    input:
        script = f"{LOC}scripts/fit_gam.R",
        rds    = f"{LOC}rds/{ITER}/original/{{distribution}}_{{n}}_{{method}}.rds"
    output:
        f"{LOC}rds/{ITER}/gam/{{distribution}}_{{n}}_{{method}}.rds"
    shell:
        "Rscript {input.script} --n {wildcards.n} --method {wildcards.method} "
        "--loc {LOC}"
```

The smoothing step reads the simulation output and writes its own file, so
re-tuning the smoother does not rerun the simulation. Every stage you can afford to
separate is a stage you will not have to repeat --- and in a project where the
first stage takes a day, that is the difference between iterating on a figure in
minutes and iterating on it overnight.

## Optional expensive pieces

Some comparisons are too slow to run by default. Make them a switch:

```{.yaml filename="config.yaml"}
desparsified: false   # adds hours; needed only for the final version
```

```{.python filename="Snakefile"}
DESPARSIFIED = config.get("desparsified", False)

rule figure5:
    input:
        rds = expand(
            f"{LOC}rds/{ITER}/gam/laplace_{{method}}.rds",
            method = ["pipep", "selectiveinferenceS"] +
                     (["desparsified"] if DESPARSIFIED else [])
        )
```

The list of inputs is computed by Python, so a flag in the config changes which jobs
exist at all. The figure script gets the same flag and adjusts its legend
accordingly. Someone reproducing your results gets something in an hour by default,
and everything if they are willing to wait.

## What this buys you

A `README` for a project of this shape can honestly say:

> With Snakemake installed, `snakemake --cores=1` runs the simulations and builds
> the manuscript. Start with `snakemake figure1` to confirm your setup works.

Which is a strong claim, and it is checkable --- by a reviewer, by a reader, and by
you in eighteen months when someone asks where a number came from.

## Where to go next {#sec-next}

Everything in this book runs locally. Three directions to read about when you
outgrow that:

**Clusters.** Snakemake can submit each job to SLURM or SGE instead of running it
locally. This is a *profile* --- a directory containing a `config.yaml` with the
submit command and resource defaults --- plus an executor plugin
(`snakemake-executor-plugin-slurm`, or `-cluster-generic` for SGE). The Snakefile
does not change; you add `--profile slurm` to the command line. Per-rule
`resources: mem_mb=..., runtime=...` become the memory and walltime requests. Note
that this changed substantially in Snakemake 8, so check the version of anything
you read. The [Iowa HPC tutorial](https://iowabiostat.github.io/hpc/12.html) has a
worked SGE example.

**Containers.** A `container:` directive at the top of the Snakefile runs every job
inside a Docker or Apptainer image, which pins your R version and every package
along with it. This is the piece that turns "reproducible on my machine" into
"reproducible".

**Checkpoints.** Everything here assumes the DAG is known before anything runs.
When the set of jobs depends on a result --- fit a model, then run one job per
selected variable --- you need `checkpoint` instead of `rule`, which lets Snakemake
re-plan mid-run.

The [Snakemake documentation](https://snakemake.readthedocs.io/) covers all three,
and the [best practices](https://snakemake.readthedocs.io/en/stable/snakefiles/best_practices.html)
page is worth a read once the ideas here are comfortable.
